Two long-axis dimensions of hippocampal-cortical integration support memory function across the adult lifespan

Version 1

Abstract

Abstract

The hippocampus is a complex structure critically involved in numerous behavior-regulating systems. In young adults, multiple overlapping spatial modes along its longitudinal and transverse axes describe the organization of its functional integration with neocortex, extending the traditional framework emphasizing functional differences between sharply segregated hippocampal subregions. Yet, it remains unknown whether these modes (i.e., gradients) persist across the adult human lifespan, and relate to memory and molecular markers associated with brain function and cognition. In two independent samples, we demonstrate that the principal anteroposterior and second-order, mid-to-anterior/posterior hippocampal modes of neocortical functional connectivity, representing distinct dimensions of macroscale cortical organization, manifest across the adult lifespan. Specifically, individual differences in topography of the second-order gradient predicted episodic memory and mirrored dopamine D1 receptor distribution, capturing shared functional and molecular organization. Older age was associated with less distinct transitions along gradients (i.e., increased functional homogeneity). Importantly, a youth-like gradient profile predicted preserved episodic memory – emphasizing age-related gradient dedifferentiation as a marker of cognitive decline. Our results underscore a critical role of mapping multidimensional hippocampal organization in understanding the neural circuits that support memory across the adult lifespan.

The hippocampus plays a critical role in human behavior beyond its well-established involvement in memory and spatial navigation1–3. Contemporary views hold that its broad involvement in cognition emerges through the combination of its intrinsic circuitry and its widespread neocortical connections – placing it at the interface of multiple behavioral systems2,4. Characterizing organizational principles of hippocampal integration with the larger neocortical landscape is therefore key to our understanding of its contribution to cognition and to the many diseases associated with its dysfunction5–8.

Animal models9,10, together with histological and functional descriptions in humans11–13, emphasize the hippocampus transverse (mediolateral) and longitudinal (anteroposterior) axes in determining its functional organization, contribution to behavior14,15, and vulnerability to neurological disease16,17. In humans, functional analogues of the hippocampus canonical internal circuitry and its anatomical connections with neocortical areas have successfully been provided by resting-state functional magnetic resonance imaging (fMRI)12,18, at a coarse scale confirming the anteroposterior differentiation in connectivity observed in the animal literature9,19,20. Despite the consistency by which this anteroposterior organizational dimension emerges21, significant questions however remain regarding its spatial distribution across cortex and its contribution to behavior across the adult human lifespan.

Lack of consensus is especially evident in terms of the hippocampus functional connectivity with the default-mode network (DMN), encompassing core areas of the brain’s system for memory-guided behavior4,22. Several studies primarily attribute integration with the DMN to the posterior hippocampus4,12,23,24, consistent with its anatomical connections to midline posterior parietal areas19. Other sources emphasize the anterior hippocampus as driving connectivity with the DMN25–27, on the basis of its anatomical connections with ventromedial prefrontal areas19. These inconsistencies ultimately limit our understanding of hippocampal functional specialization, and the impact on cognition of the heterogeneous vulnerability to aging and disease observed along the hippocampus longitudinal axis17,28. We propose that inconsistencies in part stem from overlooking multiple overlapping and complementary functional modes, not discernable through traditional parcellation-based approaches, which assume homogeneous function within distinct, pre-defined, portions of the hippocampus.

Accumulating evidence in young adults indicate that hippocampal neocortical connectivity is indeed fundamentally multidimensional – organized in several overlapping, but distinct, spatial modes21,29–32. Whereas such gradient-based observations describe a principal anteroposterior mode of neocortical connectivity, they also identify orthogonal modes of long-axis and transverse variation in connectivity29–31. Importantly, these hippocampal gradients appear to reflect well-known gradients of macroscale brain function31, which express functional differentiation across distinct cortical hierarchies33,34. The principal anteroposterior gradient has been linked to functional differentiation along a task-negative/task-positive cortical dimension35, separating neural communities involved in the formation of representations from sensory input (e.g., visual, somatosensory, and DMN areas) and those involved in the modulation of these representations (e.g., frontoparietal areas of attention and control)31,34. In parallel, a secondary, non-linear, long-axis gradient is suggested to correspond to the principal unimodal-transmodal gradient of cortical function (oppositely anchored in associative DMN areas and in unimodal sensory and motor cortices33,34)31. However, the biological underpinnings of this secondary gradient are still unknown, in contrast to the principal gradient, demonstrated also in microstructure23, gray matter covariance13,36, and gene expression37.

Differences in the topography (i.e., the spatial layout) of the principal, anteroposterior, hippocampal gradient may predict episodic memory30, but current findings are restricted to younger age – and to mainly the same sample (i.e., the Human Connectome Project38)29–31. Moreover, it remains unknown in which capacity the secondary long-axis gradient contributes to behavior. Yet, a comprehensive model of the spatial properties of hippocampal functional organization should consider that hippocampal functional alterations occur across the adult lifespan, and manifest across the hippocampus in a heterogeneous manner. Specifically, aging has been linked to differential functional isolation of anterior and posterior hippocampal regions from prefrontal areas and large-scale brain networks including the DMN39,40. Such neocortical disconnection of hippocampal subregions has been linked to dysfunction during memory encoding and retrieval40–42, and in turn, to episodic memory decline42. Findings which link hippocampal isolation to increased functional homogeneity within the region, suggest that its disconnection from neocortex is driven by the spatial patterns in which Alzheimer’s disease (AD) pathology accumulates in cognitively healthy older adults43,44. Importantly, loss of functional segregation between task-negative and task-positive poles is a functional hallmark of both healthy aging45 and AD46 – introducing overall ambiguity as to whether hippocampal gradients established in young adults persist into older age.

Dopamine (DA) is one of the most important modulators of hippocampus-dependent function47,48, and influences the brain’s functional architecture through enhancing specificity of neuronal signaling49. Consistently, there is a DA-dependent aspect of maintained functional network segregation in aging which supports cognition50. Animal models suggest heterogeneous patterns of DA innervation51,52 and postsynaptic DA receptors53, across both transverse and longitudinal hippocampal axes, likely allowing for separation between DA modulation of distinct hippocampus-dependent behaviors47. Moreover, the human hippocampus has been linked to distinct DA circuits on the basis of long-axis variation in functional connectivity with midbrain and striatal regions54,55. We therefore tested the hypothesis that the organization of hippocampal neocortical connectivity partly reflects the underlying distribution of hippocampal DA receptors. Based on recent findings revealing a unimodal-transmodal organization of the most abundantly expressed DA receptor subtype, D1 (D1DR), across cortex56, we predicted predominant spatial correspondence for any hippocampal gradient conveying a unimodal-transmodal pattern across cortex.

Here, we characterize the multidimensional functional organization of the hippocampus in two independent adult-lifespan samples, and map individual differences in fine-scale topographic properties of connectivity gradients onto behavioral and molecular phenotypes. We report three hippocampal gradients displaying distinct correspondence to a) canonical gradients of cortical function, b) the organization of hippocampal DA receptors, and c) individual differences in memory function. Multivariate, data-driven, classification on gradient topography identified older adults exhibiting a youth-like gradient profile and superior memory function as distinct from age-matched older counterparts, emphasizing a behavioral significance of preserved functional hippocampal topography in older age.

Results

Multiple dimensions of hippocampal-neocortical integration across the adult lifespan

Connectopic mapping57 was applied to resting-state fMRI data (n=180, 90 men/90 women; 20-79 years; mean age = 49.8±17.4) from the DyNAMiC study58. For replication, we used an independent sample of 224 adults (122 men/102 women; 29-85 years mean age = 65.0±13.0) from the Betula project59. Connectopic mapping was used to extract the dominant modes of functional cortical connectivity within the hippocampus based on non-linear manifold learning (Laplacian eigenmaps) applied to a similarity matrix derived from connectivity fingerprints computed between each hippocampal voxel and each voxel within neocortex. This identified a set of orthogonal connectopic maps (i.e., eigenvectors) describing overlapping connectivity topographies (i.e., gradients) within the hippocampus. Gradients were computed at subject level, and at group level across the sample, separately for the left and right hippocampus. We analyzed the first three gradients, together explaining 63% and 71% of the variance in left and right-hemispheres, respectively. This number corresponded to a clear elbow in the scree plot (Supplementary Figure 1).

The principal gradient (G1), explaining 44% and 53% of the variance in left and right hemispheres, was organized along the hippocampus longitudinal axis, conveying gradual anterior-to-posterior variation in cortical connectivity (Figure 1A). This pattern of connectivity change is illustrated by dividing subject-level G1 connectopic maps into 23 long-axis bins of ∼2mm and plotting the average gradient values as a function of their distance from the most anterior hippocampal voxel30 (Figure 1B). The second-order gradient (G2), explaining 11% of the variance in both hemispheres, expressed a secondary long-axis gradient organized from the middle hippocampus towards anterior and posterior ends (Figure 1A-B). Finally, the third-order gradient (G3: explaining 8% and 7% of the variance), reflected variation along the hippocampus transverse axis, such that inferior-lateral parts of the hippocampus were separated from medial-superior parts (Figure 1A). This pattern was most pronounced in the anterior hippocampus. Inspecting G3 across sample-specific segmentations of cornu ammonis (CA1-3), dentate gyrus (DG/CA4), and subiculum subfields suggested that while CA1-3 expressed the full extent of the gradient, and DG/CA4 variation around its center, the subiculum expressed only the most inferior section of the gradient (Supplementary Figure 2).

The three gradients reflected gradients identified in young adults29–31, and were highly reproducible in the independent replication data set (Supplementary Figure 3). Correspondence between samples was determined by spatial correlations (left hemisphere: G1: r = 0.990, p < 0.001; G2: r = 0.946, p < 0.001; G3: r = 0.918, p < 0.001; right hemisphere: G1: r = 0.996, p < 0.001; G2: r = 0.969, p < 0.001; G3: r = 0.897, p < 0.001). Furthermore, the reliability of connectopic mapping to produce functional connectivity gradients was determined across varying degrees of spatial smoothing, and contrasted against connectopic maps derived from random data60. Results confirmed high stability of resting-state gradients and their efficacy in capturing inter-individual differences, whereas random data failed to produce meaningful gradients (Supplementary Figure 4).

Topographic gradients of hippocampal cortical connectivity.

A) The first three hippocampal connectopic maps (G1-G3), explaining 67% of the variance across left and right hemispheres. Similar colors convey similar patterns of cortical connectivity. Values range between 0 (blue) and 1 (yellow). B) Plots convey change in connectivity along the anteroposterior hippocampal axis. Mean values from 23 hippocampal bins (each ∼2mm) are plotted against their distance (in mm) from the most anterior hippocampal voxel. Values were estimated based on subject-level gradients and averaged across participants. G1 conveys gradual change in connectivity along an anteroposterior gradient. G2 conveys gradual change in connectivity along a second-order long-axis gradient, separating the middle hippocampus from anterior and posterior ends. G3 conveys close to no change in connectivity along the longitudinal axis, with connectivity change instead organized in a primarily medial-lateral gradient. C) Cortical projections for G1, G2, and G3. Values range between 0 (blue) and 1 (yellow). D) The order of cortical networks in gradient space. Density plots visualize the distribution of gradient values for seven cortical networks (Yeo et al., 2011). E) Correlations between cortical patterns of hippocampal gradients and the three primary gradients of cortical functional organization, which are exemplified at the top of each graph (Margulies et al., 2016).

Hippocampal gradients reflect distinct dimensions of macroscale cortical organization

The projection of G1 onto cortex conveyed a pattern linking medial orbitofrontal, temporolimbic, and medial parietal regions at the anterior end of the gradient with occipital and frontoparietal regions at the posterior end (Figure 1C). For further characterization, we computed G1 gradient values within seven cortical networks61, and examined their position in gradient space. This placed the DMN, limbic, and somatomotor networks at anterior-to-middle parts of the gradient, whereas visual, ventral attention, and frontoparietal networks toward the posterior end of the gradient (Figure 1D). In contrast, G2 exhibited a unimodal-transmodal pattern across cortex, linking the middle hippocampus to frontal and posterior parietal regions, and anterior and posterior hippocampal ends to somatomotor and occipital regions (Figure 1C). Consistently, the DMN and frontoparietal network mapped onto G2 at one end, and visual and somatomotor networks at the other (Figure 1D). Across cortex, G3 primarily separated temporal and insular areas from medial parietal and medial frontal areas (Figure 1C). Aligning with cortical connectivity previously reported for hippocampal subfields29,62, areas of the DMN most strongly mapped onto the most inferior end of G3, consistent with the connectivity profile of the subiculum, whereas ventral attention and somatomotor networks had a stronger medial position along G3, aligning with reported connectivity of CA1-3 (Figure 1D).

Next, we linked the cortical patterns of gradients to the established macroscale layout of cortical function34 using Spearman’s rank correlation analyses. G1 showed significant correspondence with the cortical representation-modulation gradient34, differentiating the task-negative DMN and somatomotor networks from task-positive areas of attention and control (left G1: Spearman’s r = -0.353, pspin < 0.001; right G1: Spearman’s r = -0.406, pspin < 0.001; Figure 1E). G1’s correlation with this cortical gradient was greater than its correlations to other cortical gradients (6.3 < Z < 9.6, ps < 0.001), as well as significantly stronger than correlations observed for G2 and G3 (1.6 < Z < 6.5, ps < 0.05). In contrast, G2 corresponded to the principal unimodal-transmodal gradient of cortical function (left G2: Spearman’s r = 0.502, pspin < 0.001; right G2: Spearman’s r = 0.536, pspin < 0.001; Figure 1E), to a greater extent than G1 and G3 (4.9 < Z < 10.9, ps < 0.001) and in comparison to other cortical gradients (5.2 < Z < 9.4, ps < 0.001). Finally, the cortical pattern of G3 showed overall weaker correspondence with the canonical cortical gradients. G3’s correlations with the unimodal-transmodal gradient (left G3: Spearman’s r = -0.32, pspin = 0.011), and the representation-modulation gradient (right G3: Spearman’s r = 0.191, pspin = 0.009), were significantly weaker compared their counterparts of G2 (Z(998) = 4.9, p < 0.001) and G1 (Z(998) = 5.3, p < 0.001). These multiple lines of evidence contribute to a model of hippocampal functional organization across the adult human lifespan in which G1 and G2 constitute local representations of distinct macroscale cortical motifs.

Distinct patterns of behavioral transitions along G1 and G2

Given the correspondence of G1 and G2 to distinct gradients of cortical function, we characterized their relevance for hippocampal functional specialization by mapping transitions in behavioral domains onto G1 and G2 using meta-analytical decoding in Neurosynth63. Correlations were assessed between meta-analytical maps of behavioral terms and twenty-percentile bins of each gradient’s cortical projection (Figure 2). First, a selection of terms commonly linked to anteroposterior hippocampal functional specialization13,14 were assessed across G1 and ranked based on their location along the gradient (Figure 2A). The strongest anterior loadings on G1 belonged to terms including words, social, and dementia, whereas terms of navigation, episodic memory, encoding, and recollection showed preferential posterior loadings. In contrast, behavioral transitions along G2 were expected to reflect its unimodal-transmodal organization (Figure 1B). To this end, terms were selected and ordered based on a previous report demonstrating unimodal-transmodal behavioral transitions across cortex34. G2 expressed a clear separation between sensorimotor and visual terms at one end from social, self-referential, and default terms at the other (Figure 2B), confirming its unimodal-transmodal organization.

Behavioral profiling of G1 and G2 across cortex.

For each gradient, columns represent twenty-percentile bins of the gradient’s cortical projection. Color shadings represent the strength of correlations between gradient bins and meta-analytical maps in Neurosynth. A) Terms commonly linked to anteroposterior hippocampal functional specialization were assessed across G1 and ranked based on their location along the gradient. B) For G2, terms were selected and ordered as to reflect a unimodal-transmodal cortical axis34. C) The correspondence between G2 and behavioral terms commonly linked to anteroposterior hippocampal functional specialization. D) The correspondence between G1 and behavioral terms expressing a unimodal-transmodal axis.

Topography of G2 reflects distribution of hippocampal dopamine D1 receptors

Next, we tested the spatial correspondence between functional gradients and the distribution of hippocampal DA D1 receptors (D1DR) measured by [11C]SCH23390 PET in the DyNAMiC58 sample. To estimate individual differences in gradients’ spatial layout, trend surface modelling (TSM) was applied to each subject-level connectopic map30,57,64. This spatial-statistics approach parameterizes gradients at subject level, yielding a set of model parameters describing the topographic characteristics of each gradient in x, y, z directions (see Methods, and Supplementary Figure 5 for model selection). Unlike voxel-wise statistical inference on gradients, which overlooks the high interdependence between voxels’ gradient values and demands rigorous correction for multiple comparisons, TSM allows for statistical inference across a region’s internal heterogeneity using a concise set of independent parameters64. Moreover, by adjusting the number of polynomial terms, TSM facilitates examination of spatial trends across gradients at coarser-to-finer levels64.

Individual maps of D1DR binding potential (BP) were also submitted to TSM, yielding a set of spatial model parameters describing the topographic characteristics of hippocampal D1DR distribution for each participant. D1DR parameters were subsequently used as predictors of gradient parameters in a multivariate GLM (controlled for age, sex, and mean FD). Individual differences in D1DR topography significantly explained topography of right-hemisphere G2 (F = 1.207, p = 0.041; partial η2 = 0.118), but not of G1 nor G3 (F 0.902-1.108, p 0.222-0.750) (Figure 3). This association was robust across multiple TSM model orders (Supplementary Figure 6). Further characterization of G2 as reflecting DA organization was conducted by utilizing additional DA markers at group-level. First, a map of D1DR distribution was formed by averaging the [11C]SCH23390 BP images across DyNAMiC participants (n=176), and a map of D2DR distribution was formed by averaging [11C]raclopride BP images in a subsample of DyNAMiC participants (n = 20). Previously published maps of DAT65 and FDOPA (https://www.nitrc.org/projects/spmtemplates) were also analyzed. Correlations across group-level TSM parameters66, revealed significant positive associations between G2 and D1, DAT, and FDOPA (D1: r = 0.501, p < 0.01; DAT: r = 0.378, p < 0.01; FDOPA: r = 0.584, p < 0.01; Figure 3C), although not D2 (D2: r = 0.131, p = 0.44), whereas correlations were not significant for G1 or G3, indicating that G2 best captures shared principles of functional and molecular organization.

Shared functional and molecular organization within the hippocampus.

A) Multivariate effects of hippocampal D1DR TSM parameters as predictors of G2 TSM parameters. Images show average organization of G2 and D1DR in the right hemisphere. Note that the arbitrary color scale of G2 has been flipped. B) Correlations between group-level TSM parameters of functional gradients and DA markers.

Dedifferentiated gradient topography in older age

Effects of age on gradient topography was assessed using multivariate GLMs including age as the predictor and gradient TSM parameters as dependent variables (controlling for sex and mean frame-wise displacement; FD). There was a significant effect of age on topographic characteristics of all three gradients. G1 displayed the greatest effect of age (left: F(9,150) = 5.853, p < 0.001, partial ƞ2 = 0.260; right: F(9,150) = 6.971, p < 0.001, partial ƞ2 = 0.298), followed by G2 (left: F(12,147) = 2.583, p = 0.004, partial ƞ2 = 0.174; right: F(12,145) = 2.635, p = 0.003, partial ƞ2 = 0.179), and G3 (left: F(12,147) = 1.973, p = 0.030, partial ƞ2 = 0.139; right: F(12,145) = 2.082, p = 0.021, partial ƞ2 = 0.147). To visualize effects, subject-level G1 and G2 values were plotted along the anteroposterior axis, averaged within young (20-39 years), middle-aged (40-59 years) and older (60-79 years) adults. Connectivity across G1 and G2 displayed less distinct differentiation at older age, depicted by the flatter curves in the older group (Figure 4A). The difference in gradient values between each consecutive pair of gradient bins was computed, and the locations of significant univariate contributions to the effect of age on these values were visualized along gradients (Figure 4B-C). For G1, this primarily localized effects to a middle hippocampal region extending posteriorly from just after the uncal apex (MNI y=21).

Effects of age on hippocampal gradients. A-B) Less specificity in connectivity change across G1 and G2 in older age.

Average values of subject-level gradient bins are plotted as a function of their distance from the most anterior hippocampal voxel. Separate lines mark young (20-39 years; gray), middle-aged (40-59 years, green), and older (60-79 years; blue) age groups. The flatter curves in the older group indicate less distinct change in connectivity patterns across gradients in older age. C) Localization of age effects along G1. Shaded fields indicate the position of significant age effects along G1. D) Localization of age effects along G2. Shaded fields indicate the position of significant age effects along G2.

Topography of hippocampal gradients predicts episodic memory performance

Next, we tested associations between topography of the three gradients and episodic memory. Using hierarchical multiple regression models in which age, sex, and mean FD were controlled for in a first step (M1), we entered TSM parameters of the three gradients as predictors of episodic memory in a step-wise manner. Models were assessed separately for left and right hemispheres. Memory performance was, across the sample, significantly predicted by G2 in the left hippocampus (Adj. R2 = 0.308, ΔR2 = 0.096, F= 1.842, p = 0.047) over and above covariates and topography of G1, which did not predict performance (Adj. R2 = 0.260, ΔR2 = 0.029, F= 0.695, p = 0.713), and neither did G3 (Adj. R2 = 0.276, ΔR2 = 0.027, F= 0.502, p = 0.910), Figure 5A. Memory performance was, within young adults, predicted by left-hemisphere G1 (Adj. R2 = 0.182, ΔR2 = 0.357, F= 2.672, p = 0.015), whereas neither G2 (Adj. R2 = 0.204, ΔR2 = 0.191, F= 1.098, p = 0.396) nor G3 (Adj. R2 = 0.384, ΔR2 = 0.236, F= 1.755, p = 0.132) improved the prediction (Figure 5B). No gradient predicted memory within middle-aged or older age groups (F 0.432-1.865, p 0.928-0.113; Supplementary Table 1).

Gradient topography and episodic memory.

A) Individual differences in topographic characteristics of left-hemisphere G2 significantly predicted episodic memory performance across the sample, over and above the first- and second-step models (M1: age, sex, in-scanner motion; G1 parameters). B) Topographic characteristics of G1 in the left hemisphere significantly predicted episodic memory performance in young adults, over and above M1 (age, sex, and in-scanner motion).

Youth-like gradient topography supports memory in older age

To investigate the functional role of G1 in old age, the principal and most age-sensitive gradient, we tested whether memory in older adults would be facilitated by youth-like gradient topography using two independent datasets. In DyNAMiC, we applied data-driven latent class analysis (LCA) to TSM parameters (residualized to account for age, sex, and mean FD) of left-hemisphere G1, which predicted episodic memory performance in younger adults. LCA yielded a two-class solution identifying two subgroups (n=19 vs. n=30) of older adults (60-79 years), which by definition differed in left-hemisphere G1 characteristics (F(9,37) = 13.778, p < 0.001, partial η2 = 0.770). A difference was also evident in the right hemisphere (F(9,37) = 3.790, p = 0.002, partial η2 = 0.480).

Individuals in the smaller subgroup were determined as exhibiting an aged gradient profile, whereas older adults in the larger subgroup as exhibiting a youth-like gradient profile. The classification based on G1 parameters extended across all three gradients in both hemispheres (Figure 6A), such that the smaller subgroup displayed marked differences from younger adults across all gradients (left G1: F(9,63) = 15.549, p < 0.001, partial η2 = 0.690; right G1: F(9,63) = 5.322, p < 0.001, partial η2 = 0.432; left G2: F(12,60) = 3.991, p < 0.001, partial η2 = 0.444; right G2: F(12,60) = 2.192, p = 0.023, partial η2 = 0.305; left G3: F(12,60) = 2.832, p = 0.004, partial η2 = 0.362; right G3: F(12,60) = 1.844, p = 0.061, partial η2 = 0.269), while the larger subgroup differed less from young adults in terms of G1 (left G1: F(9,74) = 4.416, p < 0.001, partial η2 = 0.349; right G1: F(9,74) = 3.086, p = 0.003, partial η2 = 0.273), and displayed second- and third-order gradients comparable to those in younger age (left G2: F(12,71) = 1.616, p = 0.107, partial η2 = 0.215; right G2: F(12,71) = 1.442, p = 0.168, partial η2 = 0.196; left G3: F(12,71) = 1.122, p = 0.357, partial η2 = 0.159; right G3: F(12,71) = 1.596, p = 0.112, partial η2 = 0.212). See Supplementary Figure 7 for classification results based on right-hemisphere G1.

Plotting connectivity change along G1 and G2 in the aged and youth-like subgroups revealed that the diminished topographic specificity observed across gradients in older individuals (Figure 4A) was driven by older adults with an aged gradient profile (Figure 6B). Both older subgroups displayed altered gradient organization across cortex (Figure 6C-D). The distribution of cortical networks in G1 space indicated a unimodal-transmodal organization in youth-like older adults, not evident in the aged older group (Figure 6D). The two groups did not differ in terms of age (aged: 70.8±6.0; youth-like: 68.4±4.7; t = 1.548, p = 0.128), sex (aged: 9 men/10 women; youth-like: 16 men/14 women; X2 = 0.166, p = 0.684), nor hippocampal gray matter (left hemisphere: aged: 4271.2 ml ±480.9; youth-like: 4246.8 ml ±269.1; t = 0.223, p = 0.824; right hemisphere: aged: 3866.2 ml ±446.3; youth-like: 3979.9 ml ±398.1; t = 0.929, p = 0.357). Subgroups showed comparable levels of average hippocampal D1DR availability (left: youth-like 0.257±0.06; aged 0.242±0.06; t = 0.796, p = 0.430; right: youth-like 0.242±0.06; aged 0.251±0.06 t = 0.296, p = 0.768), but displayed a pattern of differences in D1DR TSM parameters when compared to young adults supporting youth-like and aged profiles (youth-like subgroup vs. young: left: F(9,74) = 1.645, p = 0.118, partial η2 = 0.167; aged subgroup vs. young: left: F(9,62) = 3.478, p = 0.002, partial η2 = 0.335; Figure 6E).

In line with our hypothesis, we observed superior memory in older adults exhibiting a youth-like gradient profile (Figure 6E): at trend-level for the composite episodic measure (aged: 43.2±3.7; youth-like: 46.5±6.6; t = 1.958, p = 0.056), driven by a significant group difference on its word recall sub test (aged: 40.9±4.5; youth-like: 43.4±6.8; t = 2.600, p = 0.012). Word recall performance was furthermore predicted by left-hemisphere G1 parameters (over and above age, sex, and mean FD) in the youth-like older adults (Adj. R2 = 0.464, ΔR2 = 0.543, F = 3.043, p = 0.028), while no association was observed in the aged older subgroup (Adj. R2 = 0.063, ΔR2 = 0.533, F = 1.004, p = 0.518). Crucially, the ability of left G1 topography to inform classification of older adults into mnemonically distinct subgroups was replicated in the Betula sample (Supplementary Information; Figure 7C).

Gradient profiles in older age.

A) Two older subgroups were identified based on left-hemisphere G1. The first group (n=19) displayed gradient characteristics significantly different from those in young adults, whereas the second group (n=30) displayed gradient characteristics more similar to those in young adults. Bars represent comparisons of gradient TSM parameters between older subgroups and younger adults. B) Average gradient values across participants within subgroups, plotted against the distance (in mm) from the most anterior hippocampal voxel. The flatter curves in older adults with an aged gradient profile suggest less distinct change in connectivity across gradients. C) Group-level G1 and G2 for young, youth-like, and aged groups. D) Position of cortical networks in gradient space. E) Older subgroups were comparable in terms of age, sex, hippocampal gray matter volume, and average levels of hippocampal D1DR availability, while older adults with a youth-like gradient profile exhibited a more youth-like profile also in D1DR distribution, and superior episodic memory.

Discussion

In this study, we present a comprehensive multidimensional characterization of functional hippocampal-neocortical integration across the adult human lifespan, and methodically map topography of connectivity gradients onto behavioral and molecular phenotypes. Our findings reveal the presence of two distinct connectivity gradients distributed along the hippocampus longitudinal axis which contribute to episodic memory function across the lifespan. These observations underscore the importance of disentangling multiple dimensions of hippocampal functional organization in advancing our understanding of cortico-hippocampal systems for memory-related behavior. Moreover, we demonstrate in two independent samples, that maintaining a youth-like gradient profile in older age – characterized by preserved distinctiveness of connectivity change along gradients – supports mnemonic functioning, and that increased homogeneity of gradient topography may precede gray matter atrophy.

Despite converging evidence in young adults indicating multiple overlapping modes of functional connectivity across the hippocampus21,29–32, significant questions have remained regarding their spatial distribution across cortex, and their role in behavior. Here, connectopic mapping57 identified a principal anteroposterior gradient (G1); a second-order gradient of mid-to-anterior/posterior long-axis variation (G2); and a third-order gradient conveying variation across the hippocampus transverse axis (G3). Although we restrict our discussion of G3 given its low proportion of explained variance, we note that it mirrors patterns previously observed in both structure and function13,31,67. Consistently, cortical patterns of G3 matched cortical connectivity profiles previously reported for hippocampal subfields29,62. Our results confirm that functional data favors detection of anteroposterior hippocampal organization in contrast to that determined by its cytoarchitecture21, but indicate that higher-order connectivity modes may indeed carry coarse-scale information about subfield-determined organization.

The general organization of neocortical connectivity within the hippocampus showed stability across the lifespan (Figure 1), although clear effects of age were evident in the fine-scale topography of connectivity modes (Figure 4). Older age was associated with less distinct transitions in connectivity along G1 and G2, an effect that was exacerbated in a subgroup of older adults exhibiting an aged gradient profile – separated from older adults exhibiting youth-like gradient topography (Figure 6A-B). This finding constitutes an important addition to evidence highlighting increased homogeneity in hippocampal function in aging44,68. Current theories view this loss of specificity as a consequence of its functional isolation from neocortical areas, linked to tau-driven degeneration of the perforant pathway44,69. Notably, a consequence of age-related deterioration of this entorhinal-hippocampal pathway is impaired mnemonic functioning70. Importantly, older individuals exhibiting dedifferentiated gradient topography, but comparable hippocampal volumes, displayed less efficient episodic memory compared to older adults maintaining youth-like gradient topography (Figure 6; Supplementary Figure 7). This underscores the potential of gradient-based techniques to capture behaviorally relevant alterations in hippocampal function at a stage preceding structural decline.

Testing the correspondence across cortex of G1 and G2 to canonical gradients of cortical function provided support for G1 reflecting differentiation along a representation-modulation dimension (e.g., task-negative/task-positive35; Figure 1C-E), separating frontoparietal areas of attention and control from sensorimotor and DMN areas. Consistent with age-related loss of segregation between task-positive and task-negative poles45, also evident in numerous diseases involving hippocampal dysfunction (e.g., depression71, schizophrenia72, and Alzheimer’s disease46), we observed altered cortical organization of G1 in both older subgroups. A main difference between subgroups, however, included a unimodal-transmodal organization instead emerging in youth-like older adults, whereas not in the aged subgroup (Figure 6D). A meaningful role of this potential reorganization was indicated by an association between G1 topography and memory in the youth-like older adults only. In terms of G2, we provide support of its characterization as a local representation of the principal unimodal-transmodal cortical gradient (Figure 1C-E), widely demonstrated across functional, structural, and molecular modalities33,34,73. Taken together, our observations support a framework of cortico-hippocampal integration in which the DMN is positioned in different neural contexts: in one case at the opposite end from frontoparietal networks of attention and control, and in the other, opposite to sensorimotor and visual networks – indicating that an account of the hippocampus functional connectivity with the DMN is dependent on multiple neurofunctional hierarchies. The overlap of G1 and G2 may potentially reflect the superimposition of gradients and hubs of long-axis anatomical connections indicated in both the human and animal hippocampus20,74. Moreover, the observation that macroscale relationships between distinct cortical systems are mapped out by G1 and G2 in this manner may reflect the hippocampus primordial position in the laminar development of the cerebral cortex75, supporting the idea that hippocampal function stands, from a phylogenetic perspective, to inform general principles of brain organization21.

Notably, we discovered that G2, specifically, manifests organizational principles shared among function, behavior, and neuromodulation. Meta-analytical decoding reproduced a unimodal-associative axis across G2 (Figure 3B), and analyses in relation to the distribution of D1DRs – which similarly varies across cortex along a unimodal-transmodal axis76,77 – demonstrated topographic correspondence both at the level of individual differences and across the group. The topographic correspondence observed between G2 and several DA markers (Figure 3B), suggests that G2 may constitute a mode reflecting a dopaminergic phenotype, contributing to the currently limited understanding of its biological underpinnings.

Finally, whereas the topography of G1 predicted memory in young adults30, and informed a meaningful classification of older adults, our results emphasize a primary role of G2 in episodic memory across the adult lifespan. It is noteworthy that meta-analytical decoding of G2 primarily linked the unimodal connectivity patterns of anterior and posterior subregions to terms of episodic memory, encoding, and navigation (Figure 2C). G2’s role in memory might as such be considered in light of hippocampal integration with the visual system, by which it contributes to complex perceptual processes supporting memory78. In humans, there is evidence of direct hippocampal connections to early visual areas, with recent tractography-based work demonstrating connectivity as primarily localized to the posterior hippocampus and to a smaller region in the anterior hippocampus74.

The verbal nature of our memory tasks likely contributed to the left-lateralization of effects, yet, predominant left-hemisphere vulnerability to aging and age-related pathology should not be ruled out as a meaningful contributor to these effects79. Average hippocampal D1DR availability did not differ between older subgroups, but a tendency towards youth-like and aged functional profiles being mirrored in D1DR topography was observed (Figure 6E). However, longitudinal data is ultimately required to inform the underlying mechanisms of individual differences in hippocampal gradient topography in older age80,81. Future studies should, furthermore, assess gradient modulation by behavioral conditions and extend these method to clinical samples characterized by hippocampal dysfunction.

This study establishes behavioral relevance of two overlapping long-axis modes of hippocampal-neocortical functional connectivity, shedding light on their age-related dedifferentiation, and its impact on cognition. In sum, this study introduces a multidimensional framework for understanding hippocampal-neocortical integration and its interplay with memory and neuromodulation throughout the adult human lifespan.

Materials and Methods

This study included data from the DopamiNe, Age, connectoMe, and Cognition (DyNAMiC) study, for which the design and procedures have been described in detail elsewhere58,82. Here, we include the materials and methods relevant to the current study. DyNAMiC was approved by the Regional Ethical board and the local Radiation Safety Committee of Umeå, Sweden. All participants provided written informed consent prior to testing.

Participants

The DyNAMiC sample included 180 participants (20-79 years; mean age = 49.8±17.4; 90 men/90 women equally distributed within each decade). Individuals were randomly selected from the population register of Umeå, Sweden, and recruited via postal mail. Exclusion criteria implemented during the recruitment procedure included brain pathology, impaired cognitive functioning (Mini Mental State Examination < 26), medical conditions and treatment that could affect brain functioning and cognition (e.g. dementia, diabetes, and psychiatric diagnosis), and brain imaging contraindications (e.g. metal implants). All participants were native Swedish speakers. A total of 16 participants were excluded from connectopic mapping due to excessive in-scanner motion, leaving resting-state fMRI data for 164 participants (20-78 years; mean age = 48.7±17.3). As a replication data set, we used an independent sample of 224 cognitively healthy and native Swedish-speaking adults (122 men/102 women; 29-85 years mean age = 65.0±13.0) from the population-based Betula project, for which the design and recruitment procedures have been reported in detail elsewhere59.

Episodic memory

Episodic memory was measured using three tasks testing word recall, number-word recall and object-location recall, respectively58. In the word recall task, participants were presented with 16 Swedish concrete nouns that appeared successively on a computer screen. Each word was presented for 6 s during encoding with an inter-stimulus interval (ISI) of 1 s. Following encoding, participants reported as many words as they could recall by typing them using the keyboard. Two trials were completed, yielding a maximum score of 32. In the number-word task, participants encoded pairs of 2-digit numbers and concrete plural nouns (e.g., 46 dogs). During encoding, eight number-word pairs were presented, each displayed for 6 s, with an ISI of 1 s. Following encoding, nouns were presented again, in a re-arranged order, and participants had to report the 2-digit number associated with each presented noun (e.g. How many dogs?). This task included two trials with a total maximum score of 16. The third task was an object-location memory task. Here, participants were presented with a 6 × 6 square grid in which 12 objects were, one by one, shown at distinct locations. Each object-position pairing was displayed for 8 s, with an ISI of 1 s. Following encoding, all objects were simultaneously shown next to the grid for the participant to move them (in any order) to their correct position in the grid. If unable to recall the correct position of an object, participants had to guess and place the object in the grid to the best of their ability. Two trials of this task were completed, making the total maximum score 24.

A composite score of performances across the three tasks was calculated and used as the measure of episodic memory. For each of the three tasks, scores were summarized across the total number of trials. The three resulting sum scores were z-standardized and averaged to form one composite score of episodic memory performance (T score: mean = 50; SD = 10). Missing values were replaced by the average of the available observed scores.

Image acquisition

Brain imaging was conducted at Umeå University Hospital, Sweden. Structural and functional MRI data were acquired with a 3T Discovery MR 750 scanner (General Electric, WI, USA), using a 32-channel head coil. Positron emission tomography (PET) data were acquired with a Discovery PET/CT 690 scanner (General Electric, WI, USA).

Structural MR Imaging

Anatomical T1-weighted images were acquired with a 3D fast-spoiled gradient-echo sequence, collected as 176 slices with a thickness of 1 mm. Repetition time (TR) was 8.2 ms, echo-time (TE) = 3.2 ms, flip angle = 12°, and field of view (FOV) = 250 × 250 mm.

Functional MR Imaging

Functional MR data were collected during resting-state, with participants instructed to keep their eyes open and focus on a fixation cross during scanning. Images were acquired using a T2*-weighted single-shot echo-planar imaging (EPI) sequence, with a total of 350 volumes collected over 12 minutes. The functional time series was sampled with 37 transaxial slices, slice thickness = 3.4 mm, and 0.5 mm spacing, TR = 2000 ms, TE = 30 ms, flip angle = 80°, and FOV = 250 x 250 mm. Ten dummy scans were collected at the start of the sequence.

PET Imaging

PET was conducted in 3D mode with a Discovery PET/CT 690 (General Electric, WI, US) to assess whole-brain dopamine D1 receptor availability using the radioligand [11C]SCH23390. Scanning was done during a resting condition, with participants instructed to lay still and remain awake with their eyes open. To minimize head movement, a thermoplastic mask (Posicast®; CIVCO medical solutions; IA, US) was individually fitted for each participant, and attached to the bed surface during scanning. Following a low-dose CT scan (10 mA, 120 kV, and 0.8 s rotation time) for attenuation correction, an intravenous bolus injection with target radioactivity of 350 MBq [11C]SCH23390 was administered. The PET scan was a 60 min dynamic scan, with 6 x 10 s, 6 x 20 s, 6 x 40 s, 9 x 60 s, 22 x 120 s frames. The average radioactivity dose administered to participants was 337±27 MBq (range 205-391 MBq). Due to participant drop-out and technical issues, complete PET data was available for 177 DyNAMiC participants.

Image preprocessing

Hippocampal segmentation and volumetric assessment

Individual anatomical T1-weighted images were submitted to automated segmentation in FreeSurfer version 683. A mean image of participants’ normalized T1-weighted images was also segmented in FreeSurfer, and yielded hippocampal and cortical segmentations used as masks for connectopic mapping. Regional gray matter (GM) volume was estimated from subject-specific hippocampal segmentations, and were corrected for total intracranial volume (ICV; the sum of volumes for grey matter, white matter, and cerebrospinal fluid). Adjusted volumes were equal to the raw volume - b(ICV - mean ICV), where b is the regression slope of volume on ICV84. Automated segmentation of the hippocampus into subiculum, CA1-3, and DG/CA4 subfields was conducted in FreeSurfer using the group-average T1-weighted image, for sample-specific masks to overlay onto G3 (Supplementary Figure 2).

Functional MRI data

Resting-state fMRI data were preprocessed using Statistical Parametric Mapping (SPM12: Wellcome Trust Centre for Neuroimaging, http://www.fil.ion.ucl.ac.uk/spm/) implemented in an in-house software, DataZ. Functional images were slice-timing corrected, co-registered to the anatomical T1-images, and motion corrected, and underwent distortion correction using subject-specific B0-field maps. The functional data were subsequently co-registered to the anatomical T1-images again, temporally demeaned and linear and quadratic effects were removed. Next, a 36-parameter nuisance regression model was applied85, including mean cerebrospinal, white-matter, and whole-brain signal in addition to six motion parameters, including parameters’ squares, derivatives, and squared derivatives. To further control for in-scanner motion, the model also included a set of spike regressors, defined as binary vectors of motion-contaminated volumes exceeding a volume-to-volume root-mean-squared (RMS) displacement of 0.25 mm. A temporal high-pass filter (with a threshold of 0.009 Hz) was applied simultaneously as nuisance regression in order to not re-introduce nuisance signals. Finally, images were normalized to MNI space by Diffeomorphic Anatomical Registration using Exponentiated Lie algebra (DARTEL86) and smoothed with a 6-mm FWHM Gaussian kernel. Four individuals were excluded from the template-generation step due to non-pathological anatomical irregularities. In total, 16 participants were excluded due to displaying excessive in-scanner motion, as defined by displaying i) more than 20 volumes with >0.25 relative RMS difference in motion, and ii) greater than 0.2 average RMS across the run. On average, the relative RMS difference in motion across the sample was 0.090 (± 0.063), and the mean frame-wise displacement (FD) was 0.164 (± 0.104).

Dopamine D1 receptor availability

Preprocessing of PET data was performed in SPM12 (Wellcome Trust Centre for Neuroimaging, http://www.fil.ion.ucl.ac.uk/spm/). Binding potential relative to non-displaceable binding in a reference region (BPND; Innis et al., 2007), was used as an estimate of receptor availability (i.e. D1DR) in the hippocampus, for each participant defined using the FreeSurfer segmentation of their anatomical images. Cerebellum was used as reference region. PET images were corrected for head movement by using frame-to-frame image co-registration, and co-registered with T1-weighted MRI images with re-slicing to T1 voxel size. The simplified reference-tissue model (SRTM) was used to model regional time-activity course (TAC) data. Regional TAC data were adjusted for partial volume effects (PVE) by using the symmetric geometric transfer matrix (SGTM) method implemented in FreeSurfer, and an estimated point-spread-function of 2.5 mm full-width-at-half-maximum (FWHM). We additionally used data from a publicly available database of group-averaged volumetric maps of molecular target distributions (https://github.com/netneurolab/hansen_receptors). Specifically, we downloaded previously published maps of DAT65 and FDOPA (https://www.nitrc.org/projects/spmtemplates), to investigate the spatial correspondence between functional gradients and dopaminergic target distributions.

Mapping gradients of functional connectivity

Connectopic mapping57 was run through the ConGrads toolbox57 implemented in FSL88. Mapping was conducted on both subject level and group level, for the left and right hippocampus separately, and involved two main steps. First, for every hippocampal voxel, connectivity fingerprints were computed as the Pearson correlation between the voxel-wise time-series and a singular-value decomposition (SVD) representation of all cortical voxels. In a second step, non-linear manifold learning (Laplacian eigenmaps) was applied to a matrix expressing the degree of similarity between the voxel-wise fingerprints. This yields eigenvectors, so called connectopic maps, representing modes of functional connectivity (i.e. functional gradients). Each connectopic map is then projected onto cortex, for which each vertex is color coded according to the voxel in the hippocampus it correlates the most with. Since connectopic mapping at group level involves applying Laplacian eigenmaps to a group-average similarity matrix, group level mapping across the sample was conducted using the hippocampal and cortical masks derived from the FreeSurfer segmentation of a sample-mean structural image. Mapping was specified to compute 20 gradients, and a subsequent scree plot over explained variance indicated meaningful contributions of the three first connectopic maps, together explaining 67% of the variance across hemispheres (Supplementary Figure 1).

Stability of functional gradients across levels of spatial smoothing

It has recently been suggested that the reliability of connectopic mapping to detect meaningful gradients of resting-state functional connectivity may be limited due to spatial smoothing implemented during preprocessing of fMRI data60. To determine the stability of our hippocampal gradients, we conducted a series of control analyses across varying levels of smoothing, and in contrast to connectopic maps derived through connectopic mapping on random data. These analyses are presented in the Supplementary Information, and confirmed high stability of resting-state gradients and their ability to capture inter-individual differences, whereas random data failed to produce meaningful gradients (Supplementary Figure 4).

Alignment of connectopic maps across participants

To ensure optimal alignment of identified connectopic maps across participants, we employed Procrustes alignment, based on voxel-wise correlations, to order subject-level connectopic maps according to their correspondence with a set of reference maps (i.e. gradients computed at group level across the full sample). Moreover, whereas the sign of connectopic maps is arbitrary, differences therein have an impact on the spatial model parameters describing the topographic characteristics of gradients, derived through TSM in a later step. As such, the sign of subject-level connectopic maps showing negative correlations with the corresponding group-level reference map were inversed.

Trend surface modelling

Using spatial statistics, the topography of a connectopic map can be represented by a small number of spatial model parameters. This parameterization enables analyses of inter-individual differences, and is achieved through trend surface modelling (TSM), implemented in a third step of the ConGrads analysis pipeline57. In this step, the spatial pattern of each subject-level connectopic map is approximated by estimating a spatial statistical model. Model estimation involves fitting a set of polynomial basis functions along canonical axes of the connectopic map. In MNI space, this entails estimation along x, y, and z axes of the hippocampus. Thus, fitting a polynomial of degree 1 yields three TSM parameters (x, y, z), with any increase in model order corresponding to an increase in number of parameters (e.g. 6 parameters for the second model order: x, y, z, x2, y2, z2; 9 parameters for the third model order, etc.). Trend surface models are fitted with an increasing polynomial degree using Bayesian linear regression, which provides likelihood estimates that can be used for subsequent model selection. Here, model selection was based on three information sources: a) the Bayesian Information Criterion (BIC) across subjects for models estimated at orders 1-10; b) the % explained variance in connectopic maps by each model; and c) visual inspection of group-level gradients reconstructed from TSM parameters at different model orders. The purpose of using multiple information sources, as opposed to simply BIC, was to find a trade-off between high-quality reconstructions of gradients by TSM models, while keeping the number of model parameters sufficiently low for multivariate statistical analyses. A model order of 3 (=9 TSM parameters) was selected for G1, whereas a model order of 4 (=12 TSM parameters) was selected for G2 and G3 (Supplementary Figure 5). Each gradient’s set of TSM parameters were then used as either dependent or independent variables in multivariate GLMs investigating links between gradient topography and variables such as age, episodic memory performance, and D1DR distribution.

Transitions in connectivity as a function of the hippocampal longitudinal axis

To visualize the orthogonal patterns of change in connectivity conveyed by each gradient, and to aid in the interpretation of age effects, we divided each subject-level connectopic map into 23 bins of ∼2mm along the hippocampus anterior-posterior axis and estimated the average gradient value (ranging from 0-1) for each bin. Plotting the values of each bin against their distance in mm from the most anterior voxel in the hippocampus as such demonstrates the pattern of change in connectivity along the anterior-posterior axis30.

Cortical projections and correlations with gradients of cortical function

In ConGrads57, cortical mapping is computed independently for each gradient based on the regression equation: where A corresponds to data inside (i.e., hippocampus) and B to data outside (i.e., cortical) the ROI, and X-1 is the pseudo-inverse of the corresponding eigenvector (stacked with a row of ones). For each hippocampal gradient, the volumetric MNI-registered group-level cortical projection map was resampled to the Human Connectome Project 32k_LR midthickness surface mesh, using the volume-to-surface algorithm with enclosing mapping available in Connectome Workbench v.1.5.0. Spearman correlations were computed between the cortical projection of each hippocampal gradient and the three gradients previously established as the main axes of cortical functional organization34. To reduce the degree of freedom, each surface projection was resampled by 1000-cortical parcels89. Statistical significance of correlations was assessed by spin-test permutation90, randomly rotating a spherical projection of the cortical maps 1000 times, with two-tailed statistical significance determined at a 95% confidence level.

Mapping behavioral transitions along gradients using Neurosynth

Transitions in behavioral domains were mapped onto G1 and G2 using meta-analytical decoding in Neurosynth63. We assessed two sets of behavioral terms (Figure 2), the first was a selection of terms commonly linked to anteroposterior hippocampal functional specialization13,14, and the second a selection of terms based on a previous report demonstrating behavioral transitions along a unimodal-transmodal cortical axis34. For correspondence with meta-analytical maps, we created region of interest masks by projecting the cortical surface of each gradient to the 2-mm volumetric MNI152 standard space. These volumetric images were then divided into five twenty-percentile bins and binarized. The resulting images were used as input to the Neurosynth decoder, yielding an r statistic associated with each behavioral term per section of each gradient.

Supporting information

eLife Assessment

This fundamental work demonstrates the importance of considering overlapping modes of functional organization (i.e. gradients) in the hippocampus, showing associations between with aging, dopaminergic receptor distribution and episodic memory. The evidence supporting the conclusions is solid, although some clarifications about testing procedures and a discussion of the limitations of the dopaminergic receptor mapping techniques employed should be provided along with analysis code. The work will be of broad interest to basic and clinical neuroscientists.

Reviewer #1 (Public Review):

The authors studied how hippocampal connectivity gradients across the lifespan, and how these relate to memory function and neurotransmitter distributions. They observed older age with less distinct transitions and observed an association between gradient de-differentiation and cognitive decline.

This is overall an innovative and interesting study to assess gradient alterations across the lifespan and its associations to cognition.

The paper is well-written, and the methods appear sound and thoughtful. There are several strengths, including the inclusion of two independent cohorts, the use of gradient mapping and alignment techniques, and an overall sound statistical and analysis framework. There are several areas for potential improvements in the paper, and these are listed below:

(1) The reported D1 associations appear a bit post-hoc in the current work and I was unclear why the authors specifically focussed on dopamine here, as other transmitter systems are similar present at the level of the hippocampus and implicated in aging.

Moreover, the authors may be aware that multiple PET tracers are somewhat challenged in the mesiotemporal region. Is this the case for the D1 receptor as well? The hippocampus is a small and complex structure, and PET more of a low res technique so one would want to highlight and discuss the limitations of the correlations with PET maps here and/or evaluate whether the analysis adds necessary findings to the study.

From my (perhaps somewhat biased) perspective, it might be valuable to instead or in addition look at measures of hippocampal microstructure and how these relate to the functional aging effects. This could be done, if available, using data from the same subjects (eg based on quantitative MRI contrasts and/or structural MRI) and/or using contextualization findings as implemented in eg hippomaps.readthedocs.io

(2) Can the authors clarify why they did not replicate based on cohorts that are more widely used in the community and open access, such as CamCAN and/or HCP-Aging? It might connect their results with other studies if an attempt was made to also show that findings persist in either of these repositories.

(3) The authors applied TSM and related these parameters to topographic changes in the gradients. I was wondering whether and how such an approach controls for autocorrelation present in both the PET map and gradients. Could the authors clarify?

(4) The TSM approach quantifies the gradients in terms of x/y/z direction in a cartesian coordinate system. Wouldn't a shape intrinsic coordinate system in the hippocampus also be interesting, and perhaps even be more efficient to look at here (see eg DeKraker 2022 eLife or Paquola et al 2020 eLife)?

Reviewer #2 (Public Review):

Summary:

This paper derives the first three functional gradients in the left and right hippocampus across two datasets. These gradient maps are then compared to dopamine receptor maps obtained with PET, associated with age, and linked to memory. Results reveal links between dopamine maps and gradient 2, age with gradients 1 and 2, and memory performance.

Strengths:

This paper investigates how hippocampal gradients relate to aging, memory, and dopamine receptors, which are interesting and important questions. A strength of the paper is that some of the findings were replicated in a separate sample.

Weaknesses

The paper would benefit from added clarification on the number of models/comparisons for each test. Furthermore, it would be helpful to clarify whether or not multiple comparison correction was performed and - if so - what type or - if not - to provide a justification. The manuscript would furthermore benefit from code sharing and clarifying which results did/did not replicate.

Reviewer #3 (Public Review):

Summary:

In this study, the authors analyzed the complex functional organization of the hippocampus using two separate adult lifespan datasets. They investigated how individual variations in the detailed connectivity patterns within the hippocampus relate to behavioral and molecular traits. The findings confirm three overlapping hippocampal gradients and reveal that each is linked to established functional patterns in the cortex, the arrangement of dopamine receptors within the hippocampus, and differences in memory abilities among individuals. By employing multivariate data analysis techniques, they identified older adults who display a hippocampal gradient pattern resembling that of younger individuals and exhibit better memory performance compared to their age-matched peers. This underscores the behavioral importance of maintaining a specific functional organization within the hippocampus as people age.

Strengths:

The evidence supporting the conclusions is overall compelling, based on a unique dataset, rich set of carefully unpacked results, and an in-depth data analysis. Possible confounds are carefully considered and ruled out.

Weaknesses:

No major weaknesses. The transparency of the statistical analyses could be improved by explicitly (1) stating what tests and corrections (if any) were performed, and (2) justifying the elected statistical approaches. Further, some of the findings related to the DA markers are borderline statistically significant and therefore perhaps less compelling but they line up nicely with results obtained using experimental animals and I expect the small effect sizes to be largely related to the quality and specificity of the PET data rather than the derived functional connectivity gradients.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

The authors studied how hippocampal connectivity gradients across the lifespan, and how these relate to memory function and neurotransmitter distributions. They observed older age with less distinct transitions and observed an association between gradient de-differentiation and cognitive decline.

This is overall an innovative and interesting study to assess gradient alterations across the lifespan and its associations to cognition.

The paper is well-written, and the methods appear sound and thoughtful. There are several strengths, including the inclusion of two independent cohorts, the use of gradient mapping and alignment techniques, and an overall sound statistical and analysis framework. There are several areas for potential improvements in the paper, and these are listed below:

We thank the Reviewer for their positive assessment and summary of our work. We address each of the Reviewer’s comments below, and outline the revisions we have made to the manuscript based on the Reviewer’s suggestions.

(1) The reported D1 associations appear a bit post-hoc in the current work and I was unclear why the authors specifically focussed on dopamine here, as other transmitter systems are similar present at the level of the hippocampus and implicated in aging.

Other neurotransmitter systems may indeed be relevant in the context of hippocampal function in aging. In this study, however, we included a specific research question about the DA D1 receptor (D1DR) based on previous research 1) emphasizing the role of DA neuromodulation in maintaining functional network segregation in aging to support cognition (Pedersen et al., 2023), 2) reporting heterogeneous distribution of DA markers across the hippocampus, supporting efficient modulation of distinct behaviors (Dubovyk & ManahanVaughan, 2019; Edelmann & Lessmann, 2018; Gasbarri et al., 1994; Kempadoo et al., 2016), and 3) demonstrating the spatial distribution of D1DRs as varying across neocortex along a unimodal-transmodal gradient (Pedersen et al., 2024). To which degree this variation might be reflected in cortico-hippocampal connectivity, however, remained to be investigated. As such, one of the study’s specific aims was to evaluate the spatial distribution of D1DRs as a molecular correlate of the hippocampus’ functional organization. Importantly, we were interested in mapping associations between individual differences in the organization of connectivity and D1DRs. This was uniquely enabled by utilizing the DyNAMiC sample, as it includes structural and functional MRI data in combination with D1DR PET in the same individuals across the adult lifespan (n=180). However, after observing significant spatial correspondence between functional organization and D1DR expressed by the second hippocampal gradient (G2), we did indeed perform complimentary analyses with group-averaged data of additional dopamine markers (D2DR from a subsample of our participants, as well as DAT and FDOPA from open sources) to test the generalizability of the original finding. Taken together, the original analyses based on subject-level data and complimentary group-level analyses provided support for the interpretation of G2 as a dopaminergic mode.

We have updated the manuscript to clarify the focus on the D1 receptor and the contribution of including additional DA markers.

Updated paragraph in the Introduction, pages 5-6:

“Dopamine (DA) is one of the most important modulators of hippocampus-dependent function(47,48), and influences the brain’s functional architecture through enhancing specificity of neuronal signaling(49). Consistently, there is a DA-dependent aspect of maintained functional network segregation in aging which supports cognition(50). Animal models suggest heterogeneous patterns of DA innervation(51,52) and postsynaptic DA receptors(53), across both transverse and longitudinal hippocampal axes, likely allowing for separation between DA modulation of distinct hippocampus-dependent behaviors(47). Moreover, the human hippocampus has been linked to distinct DA circuits on the basis of long-axis variation in functional connectivity with midbrain and striatal regions(54,55). Taken together with recent findings revealing a unimodal-transmodal organization of the most abundantly expressed DA receptor subtype, D1 (D1DR), across cortex(56), we tested the hypothesis that the organization of hippocampal-neocortical connectivity partly reflects the underlying distribution of hippocampal DA receptors, predicting predominant spatial correspondence for any hippocampal gradient conveying a unimodal-transmodal pattern across cortex.”

Updated sections in the Results, page 13-14:

“Our next aim was to investigate to which extent the distribution of hippocampal DA D1 receptors (D1DRs), measured by [11C]SCH23390 PET in the DyNAMiC(58) sample, may serve as a molecular correlate of the hippocampus’ functional organization.”

“Complimentary analyses were then conducted to further evaluate G2 as a dopaminergic hippocampal mode by utilizing additional DA markers at group-level.”

Moreover, the authors may be aware that multiple PET tracers are somewhat challenged in the mesiotemporal region. Is this the case for the D1 receptor as well? The hippocampus is a small and complex structure, and PET more of a low res technique so one would want to highlight and discuss the limitations of the correlations with PET maps here and/or evaluate whether the analysis adds necessary findings to the study.

We thank the Reviewer for raising this point. The lower resolution of PET is indeed a relevant aspect to consider when quantifying D1DR availability in the hippocampus, even though previous research indicate high test-retest reliability of [11C]SCH23390 PET measurement in this region (Kaller et al., 2017). We have now elaborated on PET limitations in the Discussion of the revised manuscript.

In our study, we made efforts to reduce potential partial volume effects (PVE) by correcting our PET data, and tested spatial associations between our functional gradients and D1DR maps using trend-surface modelling (TSM), rather than through voxel-wise comparisons. This allowed us to evaluate the spatial correspondence between functional connectivity and D1DRs at a level of spatial trends, estimated using TSM models computed at increasing levels of complexity. The results showed consistent spatial overlap between G2 and D1DRs across these models, that is, across spatial trends described at coarser-to-finer scales. Furthermore, this was replicated across several DA markers with PET and SPECT data from independent samples.

Taken together, we agree with the Reviewer that the spatial correspondence observed between G2 and hippocampal D1DRs should be interpreted in the context of resolution-related limitations inherent to PET imaging. However, we strongly believe that our DA analyses offer valuable insight to the molecular underpinnings of hippocampal functional organization.

Updated paragraph in the Discussion, pages 25-26:

“We discovered that G2, specifically, manifested organizational principles shared among function, behavior, and neuromodulation. Meta-analytical decoding reproduced a unimodalassociative axis across G2 (Figure 3B), and analyses in relation to the distribution of D1DRs – which vary across cortex along a unimodal-transmodal axis(76,77) – demonstrated topographic correspondence both at the level of individual differences and across the group. It should, however, be acknowledged that PET imaging in the hippocampus is associated with resolutionrelated limitations, although previous research indicate high test-retest reliability of [11C]SCH23390 PET to quantify D1DR availability in this region(78). As such, mapping the distribution of hippocampal D1DRs at a fine spatial scale remains challenging, and replication of our results in terms of overlap with G2 is needed in independent samples. Here, we evaluated the observed spatial overlap between G2 topography and D1DRs across multiple TSM model orders, showing correspondence between modalities from simple to more complex parameterizations of their spatial properties. Topographic correspondence was additionally observed between G2 and other DA markers from independent datasets (Figure 3B), suggesting that G2 may constitute a mode reflecting a dopaminergic phenotype, which contributes to the currently limited understanding of its biological underpinnings.”

From my (perhaps somewhat biased) perspective, it might be valuable to instead or in addition look at measures of hippocampal microstructure and how these relate to the functional aging effects. This could be done, if available, using data from the same subjects (eg based on quantitative MRI contrasts and/or structural MRI) and/or using contextualization findings as implemented in eg hippomaps.readthedocs.io

We thank the Reviewer for this suggestion. We performed additional analyses investigating the spatial overlap between our connectivity gradients and estimates of hippocampal microstructure, computed as the ratio of T1- over T2-weighted (T1w/T2w) images (Glasser & Von Essen, 2011; vos de Wael et al., 2018). Analyses of spatial correspondence then followed the TSM-based method used to test the spatial overlap between functional connectivity gradients and D1DR distribution. Applying TSM to the T1w/T2w image computed for each participant yielded subject-level model parameters describing microstructure topography, which were then entered as predictors of connectivity topography in multivariate GLMs (separate models for each gradient and hemisphere, 6 models in total).

Analyses revealed that microstructure of the right hippocampus significantly predicted gradient topography of right-hemisphere G1 (F = 1.325, p = 0.034), while no other links between connectivity gradients and microstructure emerged as significant (F 0.930-1.184, ps 0.7060.079).

These results, suggesting an association along the anteroposterior axis, deviate from previous findings linking hippocampal microstructure to G3-like, medial-lateral, connectivity organization (vos de Wael et al., 2018). As we believe that comprehensive analyses of our gradients in relation to microstructure across the lifespan would be best addressed in future work, we have not included these analyses of microstructure in the revised manuscript.

(2) Can the authors clarify why they did not replicate based on cohorts that are more widely used in the community and open access, such as CamCAN and/or HCP-Aging? It might connect their results with other studies if an attempt was made to also show that findings persist in either of these repositories.

We agree with the Reviewer that replication in samples such as CamCAN and/or HCP-Aging would provide valuable opportunities to connect our findings with those of other studies using those datasets. Here, we included the Betula dataset (Nilsson et al., 2004) as our replication sample, as it was immediately available to us, included a large sample of adults in a comparable age, and a word recall episodic memory task closely aligned with the one included in DyNAMiC. Importantly, leveraging the Betula dataset as our replication sample allows us to link our findings to a wide range of previous studies central to the understanding of neurocognitive aging in general, and hippocampal aging in particular (Nyberg, 2017; Nyberg et al., 2020). Betula is a large longitudinal project that has been tracking individuals since 1988, and is part of the National E-infrastructure for Aging Research (NEAR: www.near-aging.se), through which data from several Swedish studies are made available to both national and international researchers. While we acknowledge the value of extending replication efforts to datasets like CamCAN and HCP-Aging, we emphasize the significant contribution of having replicated our connectivity gradients in the Betula dataset.

(3) The authors applied TSM and related these parameters to topographic changes in the gradients. I was wondering whether and how such an approach controls for autocorrelation present in both the PET map and gradients. Could the authors clarify?

The Reviewer raises an important topic in spatial autocorrelation. The TSM approach used to parameterize the topography of the functional gradients and D1DR distribution, and to test the spatial correspondence between modalities, did not include any specific method to control for autocorrelation. Here, we highlight two aspects of our study in relation to this point. First, we demonstrated in the Supplementary information (S. Figure 4) that autocorrelation induced by spatial smoothing likely has limited effects on overall gradient topography and the ability of TSM parameters to capture meaningful inter-individual differences in terms of age. Second, in the case of spatial overlap effects being significantly impacted by autocorrelation, we would expect the association between right-hemisphere G2 and D1DR topography to similarly emerge for G2 in the left hemisphere. The absence of such an association may speak to a limited effect of spatial autocorrelation.

(4) The TSM approach quantifies the gradients in terms of x/y/z direction in a cartesian coordinate system. Wouldn't a shape intrinsic coordinate system in the hippocampus also be interesting, and perhaps even be more efficient to look at here (see eg DeKraker 2022 eLife or Paquola et al 2020 eLife)?

This is a very relevant question and we appreciate the Reviewer’s suggestion. We recognize that there may be several benefits associated with adopting a shape-intrinsic coordinate system when characterizing effects in the hippocampus, given its curved/folded anatomy. Approaches like the ones adopted in DeKraker et al., 2022 and Paquola et al., 2020, utilizes geodesic coordinate frameworks to represent the hippocampus in surface space, enabling mapping of connectivity onto the hippocampal surface while respecting its inherent curvature and topology. We anticipate that quantifying gradients within such a framework would especially benefit identification of connectivity change across the hippocampal surface relative to reference points such as subfield boundaries, while minimizing effects of interindividual differences in hippocampal shape and folding. In our study, hippocampal gradients and their associated cortical patterns were computed in volumetric space, with TSM subsequently used to parameterize the change in connectivity along these gradients. This indeed yields a description of connectivity change within a coordinate system less specific to hippocampal anatomy, but may favor generalizability and integration with previous gradient findings within and beyond the hippocampus (e.g., Przeździk et al., 2019; Tian et al., 2020; Katsumi et al., 2023; Navarro-Schröder et al., 2015), as well as connections with broader neuroimaging frameworks through techniques such as meta-analytical decoding. In our view, the different coordinate frameworks offer complimentary insight to hippocampal organization, and while we have opted to not undertake novel analyses to explore our gradients within a geodesic coordinate system for the purposes of this paper, we recognize the importance of such evaluation of our gradients in future analyses. We have made updates to the Discussion in the revised manuscript on this topic (pages 23-24):

“Greater anatomical specificity, with more precise characterization of connectivity in relation to subfield boundaries while minimizing effects of inter-individual differences in hippocampal shape and folding, might be achieved by adopting techniques implementing a geodesic coordinate system to represent effects within the hippocampus(68,69).”

Reviewer #2 (Public Review):

Summary:

This paper derives the first three functional gradients in the left and right hippocampus across two datasets. These gradient maps are then compared to dopamine receptor maps obtained with PET, associated with age, and linked to memory. Results reveal links between dopamine maps and gradient 2, age with gradients 1 and 2, and memory performance.

Strengths:

This paper investigates how hippocampal gradients relate to aging, memory, and dopamine receptors, which are interesting and important questions. A strength of the paper is that some of the findings were replicated in a separate sample.

Weaknesses:

The paper would benefit from added clarification on the number of models/comparisons for each test. Furthermore, it would be helpful to clarify whether or not multiple comparison correction was performed and - if so - what type or - if not - to provide a justification. The manuscript would furthermore benefit from code sharing and clarifying which results did/did not replicate.

We thank the Reviewer for their positive assessment and suggestions regarding further clarifications. We have addressed the Reviewer’s comments in a point-by-point manner under the “Recommendations for the authors” section.

Reviewer #3 (Public Review):

Summary:

In this study, the authors analyzed the complex functional organization of the hippocampus using two separate adult lifespan datasets. They investigated how individual variations in the detailed connectivity patterns within the hippocampus relate to behavioral and molecular traits. The findings confirm three overlapping hippocampal gradients and reveal that each is linked to established functional patterns in the cortex, the arrangement of dopamine receptors within the hippocampus, and differences in memory abilities among individuals. By employing multivariate data analysis techniques, they identified older adults who display a hippocampal gradient pattern resembling that of younger individuals and exhibit better memory performance compared to their age-matched peers. This underscores the behavioral importance of maintaining a specific functional organization within the hippocampus as people age.

Strengths:

The evidence supporting the conclusions is overall compelling, based on a unique dataset, rich set of carefully unpacked results, and an in-depth data analysis. Possible confounds are carefully considered and ruled out.

Weaknesses:

No major weaknesses. The transparency of the statistical analyses could be improved by explicitly (1) stating what tests and corrections (if any) were performed, and (2) justifying the elected statistical approaches. Further, some of the findings related to the DA markers are borderline statistically significant and therefore perhaps less compelling but they line up nicely with results obtained using experimental animals and I expect the small effect sizes to be largely related to the quality and specificity of the PET data rather than the derived functional connectivity gradients.

We thank the Reviewer for the thoughtful summary and positive assessment of our work. To increase transparency of the statistical analyses, we have in the revised manuscript added information regarding statistical tests and corrections for multiple comparisons. In the Results, p-values were reported at an uncorrected statistical threshold, and we have in the revised manuscript included the corresponding p-values adjusted for multiple comparisons using the Benjamini-Hochberg method to control the false discovery rate (FDR). Finally, in the revised manuscript, we have now elaborated on the potential limitations of our PET analyses and we include the updated paragraph below.

Addition made to the Results section, page 13:

“Individual maps of D1DR binding potential (BP) were also submitted to TSM, yielding a set of spatial model parameters describing the topographic characteristics of hippocampal D1DR distribution for each participant. D1DR parameters were subsequently used as predictors of gradient parameters in one multivariate GLM per gradient (in total 6 GLMs, controlled for age, sex, and mean FD). Results are reported with p-values at an uncorrected statistical threshold and p-values after adjustment for multiple comparisons using the Benjamini-Hochberg method to control the false discovery rate (FDR).”

Addition made to the Results section, page 15:

“Effects of age on gradient topography were assessed using multivariate GLMs including age as the predictor and gradient TSM parameters as dependent variables (controlling for sex and mean frame-wise displacement; FD). One model was fitted per gradient and hemisphere, each model including all TSM parameters belonging to a gradient (in total, 6 GLMs).”

Addition made to the Results section, page 17:

“Models were assessed separately for left and right hemispheres, across the full sample and within age groups, yielding eight hierarchical models in total. Results are reported with p-values at an uncorrected statistical threshold and p-values after FDR adjustment.”

Updated paragraph in the Discussion, pages 25-26:

“We discovered that G2, specifically, manifested organizational principles shared among function, behavior, and neuromodulation. Meta-analytical decoding reproduced a unimodalassociative axis across G2 (Figure 3B), and analyses in relation to the distribution of D1DRs – which vary across cortex along a unimodal-transmodal axis(76,77) – demonstrated topographic correspondence both at the level of individual differences and across the group. It should, however, be acknowledged that PET imaging in the hippocampus is associated with resolutionrelated limitations, although previous research indicate high test-retest reliability of [11C]SCH23390 PET to quantify D1DR availability in this region(78). As such, mapping the distribution of hippocampal D1DRs at a fine spatial scale remains challenging, and replication of our results in terms of overlap with G2 is needed in independent samples. Here, we evaluated the observed spatial overlap between G2 topography and D1DRs across multiple TSM model orders, showing correspondence between modalities from simple to more complex parameterizations of their spatial properties. Topographic correspondence was additionally observed between G2 and other DA markers from independent datasets (Figure 3B), suggesting that G2 may constitute a mode reflecting a dopaminergic phenotype, which contributes to the currently limited understanding of its biological underpinnings.”

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Please see the comments in the public review.

We thank the Reviewer for their comments and recommendations, and have addressed them in the “Public review” section.

Reviewer #2 (Recommendations For The Authors):

(1) All statistical analyses are based on linear regressions using trend surface modeling (TSM) parameters that parameterize gradients at the subject level. These models resulted in 9 parameters for gradient 1 and 12 parameters each for gradients 2 and 3. The text states that 'Effects of age on gradient topography was assessed using multivariate GLMs including age as the predictor and gradient TSM parameters as dependent variables (controlling for sex and mean frame-wise displacement; FD)'. Please clarify whether these GLMs were fitted separately for each TSM parameter (i.e., 9+12+12=33 models for both left and right = 66 total models) or on the overall model?

We appreciate the Reviewer’s request for clarification on this matter. These GLMs were fitted on the overall TSM model, that is, through one GLM per gradient (3) and hemisphere (2), each one including all TSM parameters belonging to a gradient (in total, 6 GLMs).

In the revised manuscript, we have added more details to the Results section, page 15: “Effects of age on gradient topography were assessed using multivariate GLMs including age as the predictor and gradient TSM parameters as dependent variables (controlling for sex and mean frame-wise displacement; FD). One model was fitted per gradient and hemisphere, each model including all TSM parameters belonging to a gradient (in total, 6 GLMs).”

(2) Similarly, for memory it appears that multiple models were performed (left and right, young, middle-aged, old, whole groups). Please clarify whether and how multiple comparison correction was performed in this case.

In the revised manuscript, we have now specified the number of analyses conducted in relation to memory performance. We have also clarified that p-values were reported at an uncorrected statistical threshold, and we have in the revised manuscript included the corresponding p-values adjusted for multiple comparisons using the Benjamini-Hochberg method to control the FDR.

Updated section in the Results, page 17:

“Models were assessed separately for left and right hemispheres, across the full sample and within age groups, yielding eight hierarchical models in total. Results are reported with p-values at an uncorrected statistical threshold and p-values after FDR adjustment.”

(3) Although I applaud the authors for their replication efforts, the results do not appear to replicate well. For example, memory was linked to gradient 2 in the whole group but to gradient 1 in the young group. Furthermore, dopamine was linked to gradient 2 in the right but not the left hemisphere. Although the overall group-level gradients were very stable between the two datasets, it is not clear whether the age findings replicated and the memory subgroup findings only replicated at trend level for memory and only partially replicated at the TSM parameter level.

We thank the Reviewer for highlighting the inclusion of a replication dataset as a strength of our study, and we appreciate the recommendation to clarify to which extent results replicated. We provide a response to the Reviewer’s points below, and specify the revisions made to the manuscript in relation to this topic.

The main aim of our study was to characterize the topographic organization of functional hippocampal-neocortical connectivity within the hippocampus across the adult lifespan, as previous studies have limited their focus to younger adults. Given the lack of previous studies for comparison, together with our identification of a novel secondary long-axis connectivity gradient (G2) taking precedence over the previously established medial-lateral G3, we included the Betula sample (Nilsson et al., 2004) for the purpose of replication. There was a high level of consistency between our main dataset and our replication dataset, with gradients 1-3 in left and right hemispheres identified in both samples.

Further use of the replication dataset, beyond the identification of the connectivity gradients, was originally not planned. As such, not all subsequent analyses in the main dataset were conducted in the replication dataset. However, we found it critical to evaluate the observation that older individuals who maintained a youth-like gradient topography also exhibited higher levels of memory performance in an independent sample. This was possible given that the replication dataset included a comparable number of participants in similar ages and a word recall episodic memory task corresponding well to the one used in DyNAMiC. Overall, we conclude that these analyses replicated well across samples. Firstly, topography of lefthemisphere G1 informed the classification of older adults into youth-like and aged subgroups in both samples. Furthermore, in both samples, we observed that the older subgroups identified based on G1 topography also exhibited the youth-like vs. aged pattern in G2 topography. This pattern was, however, evident also in G3 only in the main sample, possibly suggesting a limited contribution of G3 topography in determining overall functional profiles in older age. In terms of the behavioral relevance of maintaining youth-like gradient topography in older age, we observed effects on word recall performance in both samples; although the Reviewer correctly points out that, the difference between subgroups was significant at trend-level (p = 0.058) in the replication dataset. While this indeed underscores the importance of replication efforts in additional samples, we argue that the pattern observed in our replication dataset is overall consistent with, and conveys effects in the expected direction based on, the original observations in our main dataset.

In revising the manuscript, we have performed additional analyses for replication purposes in terms of memory. Originally, we observed a significant association between G2 topography and episodic memory across the main sample. However, this effect did not remain significant after FDR adjustment for multiple comparisons. To evaluate this association further, we conducted a corresponding hierarchical multiple regression analysis in the replication dataset, which supported a role of G2 in memory (Adj. R2 = 0.368, ΔR2 = 0.081, F= 1.992, p = 0.028). Together, these analyses suggest that inter-individual differences in episodic memory performance may in part be explained by the spatial characteristics of G2 across the adult lifespan, although increased statistical power in relation to the large number of TSM parameters included in the hierarchical regression models may be needed to explore this association in smaller, age-stratified, groups. Relatedly, it is worth mentioning that higher levels of memory performance in older age were linked to the maintenance of youth-like G2 topography in both our main and replication datasets.

In parallel, topographic parameters of G1 predicted memory performance in the younger adults, which successfully replicates TSM-based results previously reported in Przeździk et al., 2019. Although similar associations were not evident within the other age groups, a link between G1 topography and memory was demonstrated in older age based on a) the identification of individuals maintaining a youth-like G1 profile and higher levels of memory, within which b) memory performance was, as in young adults, significantly predicted by G1 topography.

The spatial correspondence between G2 topography and distribution of hippocampal D1DRs was lateralized to the right, and as the Reviewer points out, as such did not replicate across hemispheres. To which extent replication across hemispheres should be expected in this case is, however, difficult to determine. Lateralization and/or hemispheric asymmetry is commonly observed in numerous hippocampal features, from the molecular level to its functional involvement in behavior (Nematis et al., 2023; Persson & Söderlund, 2015), including various dopaminergic markers tested in the animal literature (Afonso et al., 1993; Sadeghi et al., 2017). Yet, potential differences between hemispheres in D1DR availability and the spatial distribution of receptors along hippocampal axes remain less studied in humans. More data is therefore needed to determine the nature of this right-hemisphere lateralization.

In sum, we argue that our results show a good level of replication across independent datasets and across analyses in our main dataset. Whereas this study did not attempt replication of all analyses conducted in the main dataset, it has through replication across independent samples provided support for its main findings – the organization of hippocampal-neocortical connectivity along three main hippocampal gradients across the adult lifespan, and the gradient topography-based identification of older individuals maintaining a youth-like hippocampal organization in older age.

The revised manuscript includes edits made to incorporate the new analyses and clarifications of observations in relation to memory.

In the Results, page 17:

“Observing that the association between G2 and memory did not remain significant after FDR adjustment, we performed the same analysis in our replication dataset, which also included episodic memory testing. Consistent with the observation in our main dataset, G2 significantly predicted memory performance (Adj. R2 = 0.368, ΔR2 = 0.081, F= 1.992, p = 0.028) over and above covariates and topography of G1. Here, the analysis also showed that G1 topography predicted performance across the sample (Adj. R2 = 0.325, ΔR2 = 0.112, F= 3.431, p < 0.001).”

In the Discussion, page 26:

“Results linked both G1 and G2 to episodic memory, suggesting complimentary contributions of these two overlapping long-axis modes. Considered together, analyses in the main and replication datasets indicated a role of G2 topography in memory across the adult lifespan, independent of age. A similar association with G1 was only evident across the entire sample in the replication dataset, whereas results in the main sample seemed to emphasize a role of youthlike G1 topography in memory performance. In line with previous research, memory was successfully predicted by G1 topography in young adults(30), and similarly predicted by G1 in older adults exhibiting a youth-like functional profile.”

(4) Please share the data and code and add a description of data and code availability in the manuscript.

We have now made our code available, and added a statement on data and code availability in the revised manuscript.

On page 37: “Data from the DyNAMiC study are not publicly available. Access to the original data may be shared upon request from the Principal investigator, Dr. Alireza Salami. The Matlab, R, and FSL codes used for analyses included in this study are openly available at https://github.com/kristinnordin/hcgradients. Computation of gradients was done using the freely available toolbox ConGrads: https://github.com/koenhaak/congrads.”

Reviewer #3 (Recommendations For The Authors):

Please see the comments in the public review.

We thank the Reviewer for their comments and recommendations, and have addressed them in the “Public review” section.

References

Afonso, D., Santana, C., & Rodriguez, M. (1993). Neonatal lateralization of behavior and brain dopaminergic asymmetry. Brain Research Bulletin, 32(1), 11–16. https://doi.org/10.1016/0361-9230(93)90312-Y

DeKraker, J., Haast, R. A., Yousif, M. D., Karat, B., Lau, J. C., Köhler, S., & Khan, A. R. (2022). Automated hippocampal unfolding for morphometry and subfield segmentation with HippUnfold. eLife, 11, e77945. https://doi.org/10.7554/eLife.77945

Dubovyk, V., & Manahan-Vaughan, D. (2019). Gradient of expression of dopamine D2 receptors along the dorso-ventral axis of the hippocampus. Frontiers in Synaptic Neuroscience, 11. https://doi.org/10.3389/fnsyn.2019.00028

Edelmann, E., & Lessmann, V. (2018). Dopaminergic innervation and modulation of hippocampal networks. Cell and Tissue Research, 373(3), 711–727. https://doi.org/10.1007/s00441-018-2800-7

Gasbarri, A., Verney, C., Innocenzi, R., Campana, E., & Pacitti, C. (1994). Mesolimbic dopaminergic neurons innervating the hippocampal formation in the rat: A combined retrograde tracing and immunohistochemical study. Brain Research, 668(1), 71–79. https://doi.org/10.1016/0006-8993(94)90512-6

Glasser, M. F., & Essen, D. C. V. (2011). Mapping Human Cortical Areas In Vivo Based on Myelin Content as Revealed by T1- and T2-Weighted MRI. Journal of Neuroscience, 31(32), 11597–11616. https://doi.org/10.1523/JNEUROSCI.2180-11.2011

Kaller, S., Rullmann, M., Patt, M., Becker, G.-A., Luthardt, J., Girbardt, J., Meyer, P. M., Werner, P., Barthel, H., Bresch, A., Fritz, T. H., Hesse, S., & Sabri, O. (2017). Test– retest measurements of dopamine D1-type receptors using simultaneous PET/MRI imaging. European Journal of Nuclear Medicine and Molecular Imaging, 44(6), 1025–1032. https://doi.org/10.1007/s00259-017-3645-0

Katsumi, Y., Zhang, J., Chen, D., Kamona, N., Bunce, J. G., Hutchinson, J. B., Yarossi, M., Tunik, E., Dickerson, B. C., Quigley, K. S., & Barrett, L. F. (2023). Correspondence of functional connectivity gradients across human isocortex, cerebellum, and hippocampus. Communications Biology, 6(1), Article 1. https://doi.org/10.1038/s42003-023-04796-0

Kempadoo, K. A., Mosharov, E. V., Choi, S. J., Sulzer, D., & Kandel, E. R. (2016). Dopamine release from the locus coeruleus to the dorsal hippocampus promotes spatial learning and memory. Proceedings of the National Academy of Sciences, 113(51), 14835–14840. https://doi.org/10.1073/pnas.1616515114

Navarro Schröder, T., Haak, K. V., Zaragoza Jimenez, N. I., Beckmann, C. F., & Doeller, C. F. (2015). Functional topography of the human entorhinal cortex. eLife, 4, e06738. https://doi.org/10.7554/eLife.06738

Nemati, S. S., Sadeghi, L., Dehghan, G., & Sheibani, N. (2023). Lateralization of the hippocampus: A review of molecular, functional, and physiological properties in health and disease. Behavioural Brain Research, 454, 114657. https://doi.org/10.1016/j.bbr.2023.114657

Nilsson, L.-G., Adolfsson, R., Bäckman, L., Frias, C. M. de, Molander, B., & Nyberg, L. (2004). Betula: A Prospective Cohort Study on Memory, Health and Aging. Aging, Neuropsychology, and Cognition, 11(2–3), 134–148. https://doi.org/10.1080/13825580490511026

Nyberg, L. (2017). Functional brain imaging of episodic memory decline in ageing. Journal of Internal Medicine, 281(1), 65–74. https://doi.org/10.1111/joim.12533

Nyberg, L., Boraxbekk, C.-J., Sörman, D. E., Hansson, P., Herlitz, A., Kauppi, K., Ljungberg, J. K., Lövheim, H., Lundquist, A., Adolfsson, A. N., Oudin, A., Pudas, S., Rönnlund, M., Stiernstedt, M., Sundström, A., & Adolfsson, R. (2020). Biological and environmental predictors of heterogeneity in neurocognitive ageing: Evidence from Betula and other longitudinal studies. Ageing Research Reviews, 64, 101184. https://doi.org/10.1016/j.arr.2020.101184

Paquola, C., Benkarim, O., DeKraker, J., Larivière, S., Frässle, S., Royer, J., Tavakol, S.,

Valk, S., Bernasconi, A., Bernasconi, N., Khan, A., Evans, A. C., Razi, A., Smallwood, J., & Bernhardt, B. C. (2020). Convergence of cortical types and functional motifs in the human mesiotemporal lobe. eLife, 9, e60673. https://doi.org/10.7554/eLife.60673

Pedersen, R., Johansson, J., Nordin, K., Rieckmann, A., Wåhlin, A., Nyberg, L., Bäckman, L., & Salami, A. (2024). Dopamine D1-Receptor Organization Contributes to Functional Brain Architecture. Journal of Neuroscience, 44(11). https://doi.org/10.1523/JNEUROSCI.0621-23.2024

Pedersen, R., Johansson, J., & Salami, A. (2023). Dopamine D1-signaling modulates maintenance of functional network segregation in aging. Aging Brain, 3, 100079. https://doi.org/10.1016/j.nbas.2023.100079

Persson, J., & Söderlund, H. (2015). Hippocampal hemispheric and long-axis differentiation of stimulus content during episodic memory encoding and retrieval: An activation likelihood estimation meta-analysis. Hippocampus, 25(12), 1614–1631. https://doi.org/10.1002/hipo.22482

Przeździk, I., Faber, M., Fernández, G., Beckmann, C. F., & Haak, K. V. (2019). The functional organisation of the hippocampus along its long axis is gradual and predicts recollection. Cortex, 119, 324–335. https://doi.org/10.1016/j.cortex.2019.04.015

Sadeghi, L., Rizvanov, A. A., Salafutdinov, I. I., Dabirmanesh, B., Sayyah, M., Fathollahi, Y., & Khajeh, K. (2017). Hippocampal asymmetry: Differences in the left and right hippocampus proteome in the rat model of temporal lobe epilepsy. Journal of Proteomics, 154, 22–29. https://doi.org/10.1016/j.jprot.2016.11.023

Tian, Y., Margulies, D. S., Breakspear, M., & Zalesky, A. (2020). Topographic organization of the human subcortex unveiled with functional connectivity gradients. Nature Neuroscience, 1–12. https://doi.org/10.1038/s41593-020-00711-6

vos de Wael, R., Larivière, S., Caldairou, B., Hong, S.-J., Margulies, D. S., Jefferies, E., Bernasconi, A., Smallwood, J., Bernasconi, N., & Bernhardt, B. C. (2018). Anatomical and microstructural determinants of hippocampal subfield functional connectome embedding. Proceedings of the National Academy of Sciences, 115(40), 10154–10159. https://doi.org/10.1073/pnas.1803667115

Version 2

Abstract

Abstract

The hippocampus is a complex structure critically involved in numerous behavior-regulating systems. In young adults, multiple overlapping spatial modes along its longitudinal and transverse axes describe the organization of its functional integration with neocortex, extending the traditional framework emphasizing functional differences between sharply segregated hippocampal subregions. Yet, it remains unknown whether these modes (i.e., gradients) persist across the adult human lifespan, and relate to memory and molecular markers associated with brain function and cognition. In two independent samples, we demonstrate that the principal anteroposterior and second-order, mid-to-anterior/posterior hippocampal modes of neocortical functional connectivity, representing distinct dimensions of macroscale cortical organization, manifest across the adult lifespan. Specifically, individual differences in topography of the second-order gradient predicted episodic memory and mirrored dopamine D1 receptor distribution, capturing shared functional and molecular organization. Older age was associated with less distinct transitions along gradients (i.e., increased functional homogeneity). Importantly, a youth-like gradient profile predicted preserved episodic memory – emphasizing age-related gradient dedifferentiation as a marker of cognitive decline. Our results underscore a critical role of mapping multidimensional hippocampal organization in understanding the neural circuits that support memory across the adult lifespan.

The hippocampus plays a critical role in human behavior beyond its well-established involvement in memory and spatial navigation1–3. Contemporary views hold that its broad involvement in cognition emerges through the combination of its intrinsic circuitry and its widespread neocortical connections – placing it at the interface of multiple behavioral systems2,4. Characterizing organizational principles of hippocampal integration with the larger neocortical landscape is therefore key to our understanding of its contribution to cognition and to the many diseases associated with its dysfunction5–8.

Animal models9,10, together with histological and functional descriptions in humans11–13, emphasize the hippocampus transverse (mediolateral) and longitudinal (anteroposterior) axes in determining its functional organization, contribution to behavior14,15, and vulnerability to neurological disease16,17. In humans, functional analogues of the hippocampus canonical internal circuitry and its anatomical connections with neocortical areas have successfully been provided by resting-state functional magnetic resonance imaging (fMRI)12,18, at a coarse scale confirming the anteroposterior differentiation in connectivity observed in the animal literature9,19,20. Despite the consistency by which this anteroposterior organizational dimension emerges21, significant questions however remain regarding its spatial distribution across cortex and its contribution to behavior across the adult human lifespan.

Lack of consensus is especially evident in terms of the hippocampus functional connectivity with the default-mode network (DMN), encompassing core areas of the brain’s system for memory-guided behavior4,22. Several studies primarily attribute integration with the DMN to the posterior hippocampus4,12,23,24, consistent with its anatomical connections to midline posterior parietal areas19. Other sources emphasize the anterior hippocampus as driving connectivity with the DMN25–27, on the basis of its anatomical connections with ventromedial prefrontal areas19. These inconsistencies ultimately limit our understanding of hippocampal functional specialization, and the impact on cognition of the heterogeneous vulnerability to aging and disease observed along the hippocampus longitudinal axis17,28. We propose that inconsistencies in part stem from overlooking multiple overlapping and complementary functional modes, not discernable through traditional parcellation-based approaches, which assume homogeneous function within distinct, pre-defined, portions of the hippocampus.

Accumulating evidence in young adults indicate that hippocampal neocortical connectivity is indeed fundamentally multidimensional – organized in several overlapping, but distinct, spatial modes21,29–32. Whereas such gradient-based observations describe a principal anteroposterior mode of neocortical connectivity, they also identify orthogonal modes of long-axis and transverse variation in connectivity29–31. Importantly, these hippocampal gradients appear to reflect well-known gradients of macroscale brain function31, which express functional differentiation across distinct cortical hierarchies33,34. The principal anteroposterior gradient has been linked to functional differentiation along a task-negative/task-positive cortical dimension35, separating neural communities involved in the formation of representations from sensory input (e.g., visual, somatosensory, and DMN areas) and those involved in the modulation of these representations (e.g., frontoparietal areas of attention and control)31,34. In parallel, a secondary, non-linear, long-axis gradient is suggested to correspond to the principal unimodal-transmodal gradient of cortical function (oppositely anchored in associative DMN areas and in unimodal sensory and motor cortices33,34)31. However, the biological underpinnings of this secondary gradient are still unknown, in contrast to the principal gradient, demonstrated also in microstructure23, gray matter covariance13,36, and gene expression37.

Differences in the topography (i.e., the spatial layout) of the principal, anteroposterior, hippocampal gradient may predict episodic memory30, but current findings are restricted to younger age – and to mainly the same sample (i.e., the Human Connectome Project38)29–31. Moreover, it remains unknown in which capacity the secondary long-axis gradient contributes to behavior. Yet, a comprehensive model of the spatial properties of hippocampal functional organization should consider that hippocampal functional alterations occur across the adult lifespan, and manifest across the hippocampus in a heterogeneous manner. Specifically, aging has been linked to differential functional isolation of anterior and posterior hippocampal regions from prefrontal areas and large-scale brain networks including the DMN39,40. Such neocortical disconnection of hippocampal subregions has been linked to dysfunction during memory encoding and retrieval40–42, and in turn, to episodic memory decline42. Findings which link hippocampal isolation to increased functional homogeneity within the region, suggest that its disconnection from neocortex is driven by the spatial patterns in which Alzheimer’s disease (AD) pathology accumulates in cognitively healthy older adults43,44. Importantly, loss of functional segregation between task-negative and task-positive poles is a functional hallmark of both healthy aging45 and AD46 – introducing overall ambiguity as to whether hippocampal gradients established in young adults persist into older age.

Dopamine (DA) is one of the most important modulators of hippocampus-dependent function47,48, and influences the brain’s functional architecture through enhancing specificity of neuronal signaling49. Consistently, there is a DA-dependent aspect of maintained functional network segregation in aging which supports cognition50. Animal models suggest heterogeneous patterns of DA innervation51,52 and postsynaptic DA receptors53, across both transverse and longitudinal hippocampal axes, likely allowing for separation between DA modulation of distinct hippocampus-dependent behaviors47. Moreover, the human hippocampus has been linked to distinct DA circuits on the basis of long-axis variation in functional connectivity with midbrain and striatal regions54,55. Taken together with recent findings revealing a unimodal-transmodal organization of the most abundantly expressed DA receptor subtype, D1 (D1DR), across cortex56, we tested the hypothesis that the organization of hippocampal-neocortical connectivity partly reflects the underlying distribution of hippocampal DA receptors, predicting predominant spatial correspondence for any hippocampal gradient conveying a unimodal-transmodal pattern across cortex.

Here, we characterize the multidimensional functional organization of the hippocampus in two independent adult-lifespan samples, and map individual differences in fine-scale topographic properties of connectivity gradients onto behavioral and molecular phenotypes. We report three hippocampal gradients displaying distinct correspondence to a) canonical gradients of cortical function, b) the organization of hippocampal DA receptors, and c) individual differences in memory function. Multivariate, data-driven, classification on gradient topography identified older adults exhibiting a youth-like gradient profile and superior memory function as distinct from age-matched older counterparts, emphasizing a behavioral significance of preserved functional hippocampal topography in older age.

Results

Multiple dimensions of hippocampal-neocortical integration across the adult lifespan

Connectopic mapping57 was applied to resting-state fMRI data (n=180, 90 men/90 women; 20-79 years; mean age = 49.8±17.4) from the DyNAMiC study58. For replication, we used an independent sample of 224 adults (122 men/102 women; 29-85 years mean age = 65.0±13.0) from the Betula project59. Connectopic mapping was used to extract the dominant modes of functional cortical connectivity within the hippocampus based on non-linear manifold learning (Laplacian eigenmaps) applied to a similarity matrix derived from connectivity fingerprints computed between each hippocampal voxel and each voxel within neocortex. This identified a set of orthogonal connectopic maps (i.e., eigenvectors) describing overlapping connectivity topographies (i.e., gradients) within the hippocampus. Gradients were computed at subject level, and at group level across the sample, separately for the left and right hippocampus. We analyzed the first three gradients, together explaining 63% and 71% of the variance in left and right-hemispheres, respectively. This number corresponded to a clear elbow in the scree plot (Supplementary Figure 1).

The principal gradient (G1), explaining 44% and 53% of the variance in left and right hemispheres, was organized along the hippocampus longitudinal axis, conveying gradual anterior-to-posterior variation in cortical connectivity (Figure 1A). This pattern of connectivity change is illustrated by dividing subject-level G1 connectopic maps into 23 long-axis bins of ∼2mm and plotting the average gradient values as a function of their distance from the most anterior hippocampal voxel30 (Figure 1B). The second-order gradient (G2), explaining 11% of the variance in both hemispheres, expressed a secondary long-axis gradient organized from the middle hippocampus towards anterior and posterior ends (Figure 1A-B). Finally, the third-order gradient (G3: explaining 8% and 7% of the variance), reflected variation along the hippocampus transverse axis, such that inferior-lateral parts of the hippocampus were separated from medial-superior parts (Figure 1A). This pattern was most pronounced in the anterior hippocampus. Inspecting G3 across sample-specific segmentations of cornu ammonis (CA1-3), dentate gyrus (DG/CA4), and subiculum subfields suggested that while CA1-3 expressed the full extent of the gradient, and DG/CA4 variation around its center, the subiculum expressed only the most inferior section of the gradient (Supplementary Figure 2).

Topographic gradients of hippocampal cortical connectivity.

A) The first three hippocampal connectopic maps (G1-G3), explaining 67% of the variance across left and right hemispheres. Similar colors convey similar patterns of cortical connectivity. Values range between 0 (blue) and 1 (yellow). B) Plots convey change in connectivity along the anteroposterior hippocampal axis. Mean values from 23 hippocampal bins (each ∼2mm) are plotted against their distance (in mm) from the most anterior hippocampal voxel. Values were estimated based on subject-level gradients and averaged across participants. G1 conveys gradual change in connectivity along an anteroposterior gradient. G2 conveys gradual change in connectivity along a second-order long-axis gradient, separating the middle hippocampus from anterior and posterior ends. G3 conveys close to no change in connectivity along the longitudinal axis, with connectivity change instead organized in a primarily medial-lateral gradient. C) Cortical projections for G1, G2, and G3. Values range between 0 (blue) and 1 (yellow). D) The order of cortical networks in gradient space. Density plots visualize the distribution of gradient values for seven cortical networks (Yeo et al., 2011). E) Correlations between cortical patterns of hippocampal gradients and the three primary gradients of cortical functional organization, which are exemplified at the top of each graph (Margulies et al., 2016).

The three gradients reflected gradients identified in young adults29–31, and were highly reproducible in the independent replication data set (Supplementary Figure 3). Correspondence between samples was determined by spatial correlations (left hemisphere: G1: r = 0.990, p < 0.001; G2: r = 0.946, p < 0.001; G3: r = 0.918, p < 0.001; right hemisphere: G1: r = 0.996, p < 0.001; G2: r = 0.969, p < 0.001; G3: r = 0.897, p < 0.001). Furthermore, the reliability of connectopic mapping to produce functional connectivity gradients was determined across varying degrees of spatial smoothing, and contrasted against connectopic maps derived from random data60. Results confirmed high stability of resting-state gradients and their efficacy in capturing inter-individual differences, whereas random data failed to produce meaningful gradients (Supplementary Figure 4).

Hippocampal gradients reflect distinct dimensions of macroscale cortical organization

The projection of G1 onto cortex conveyed a pattern linking medial orbitofrontal, temporolimbic, and medial parietal regions at the anterior end of the gradient with occipital and frontoparietal regions at the posterior end (Figure 1C). For further characterization, we computed G1 gradient values within seven cortical networks61, and examined their position in gradient space. This placed the DMN, limbic, and somatomotor networks at anterior-to-middle parts of the gradient, whereas visual, ventral attention, and frontoparietal networks toward the posterior end of the gradient (Figure 1D). In contrast, G2 exhibited a unimodal-transmodal pattern across cortex, linking the middle hippocampus to frontal and posterior parietal regions, and anterior and posterior hippocampal ends to somatomotor and occipital regions (Figure 1C). Consistently, the DMN and frontoparietal network mapped onto G2 at one end, and visual and somatomotor networks at the other (Figure 1D). Across cortex, G3 primarily separated temporal and insular areas from medial parietal and medial frontal areas (Figure 1C). Aligning with cortical connectivity previously reported for hippocampal subfields29,62, areas of the DMN most strongly mapped onto the most inferior end of G3, consistent with the connectivity profile of the subiculum, whereas ventral attention and somatomotor networks had a stronger medial position along G3, aligning with reported connectivity of CA1-3 (Figure 1D).

Next, we linked the cortical patterns of gradients to the established macroscale layout of cortical function34 using Spearman’s rank correlation analyses. G1 showed significant correspondence with the cortical representation-modulation gradient34, differentiating the task-negative DMN and somatomotor networks from task-positive areas of attention and control (left G1: Spearman’s r = -0.353, pspin < 0.001; right G1: Spearman’s r = -0.406, pspin < 0.001; Figure 1E). G1’s correlation with this cortical gradient was greater than its correlations to other cortical gradients (6.3 < Z < 9.6, ps < 0.001), as well as significantly stronger than correlations observed for G2 and G3 (1.6 < Z < 6.5, ps < 0.05). In contrast, G2 corresponded to the principal unimodal-transmodal gradient of cortical function (left G2: Spearman’s r = 0.502, pspin < 0.001; right G2: Spearman’s r = 0.536, pspin < 0.001; Figure 1E), to a greater extent than G1 and G3 (4.9 < Z < 10.9, ps < 0.001) and in comparison to other cortical gradients (5.2 < Z < 9.4, ps < 0.001). Finally, the cortical pattern of G3 showed overall weaker correspondence with the canonical cortical gradients. G3’s correlations with the unimodal-transmodal gradient (left G3: Spearman’s r = -0.32, pspin = 0.011), and the representation-modulation gradient (right G3: Spearman’s r = 0.191, pspin = 0.009), were significantly weaker compared their counterparts of G2 (Z(998) = 4.9, p < 0.001) and G1 (Z(998) = 5.3, p < 0.001). These multiple lines of evidence contribute to a model of hippocampal functional organization across the adult human lifespan in which G1 and G2 constitute local representations of distinct macroscale cortical motifs.

Distinct patterns of behavioral transitions along G1 and G2

Given the correspondence of G1 and G2 to distinct gradients of cortical function, we characterized their relevance for hippocampal functional specialization by mapping transitions in behavioral domains onto G1 and G2 using meta-analytical decoding in Neurosynth63. Correlations were assessed between meta-analytical maps of behavioral terms and twenty-percentile bins of each gradient’s cortical projection (Figure 2). First, a selection of terms commonly linked to anteroposterior hippocampal functional specialization13,14 were assessed across G1 and ranked based on their location along the gradient (Figure 2A). The strongest anterior loadings on G1 belonged to terms including words, social, and dementia, whereas terms of navigation, episodic memory, encoding, and recollection showed preferential posterior loadings. In contrast, behavioral transitions along G2 were expected to reflect its unimodal-transmodal organization (Figure 1B). To this end, terms were selected and ordered based on a previous report demonstrating unimodal-transmodal behavioral transitions across cortex34. G2 expressed a clear separation between sensorimotor and visual terms at one end from social, self-referential, and default terms at the other (Figure 2B), confirming its unimodal-transmodal organization.

Behavioral profiling of G1 and G2 across cortex.

For each gradient, columns represent twenty-percentile bins of the gradient’s cortical projection. Color shadings represent the strength of correlations between gradient bins and meta-analytical maps in Neurosynth. A) Terms commonly linked to anteroposterior hippocampal functional specialization were assessed across G1 and ranked based on their location along the gradient. B) For G2, terms were selected and ordered as to reflect a unimodal-transmodal cortical axis34. C) The correspondence between G2 and behavioral terms commonly linked to anteroposterior hippocampal functional specialization. D) The correspondence between G1 and behavioral terms expressing a unimodal-transmodal axis.

Topography of G2 reflects distribution of hippocampal dopamine D1 receptors

Our next aim was to investigate to which extent the distribution of hippocampal DA D1 receptors (D1DRs), measured by [11C]SCH23390 PET in the DyNAMiC58 sample, may serve as a molecular correlate of the hippocampus’ functional organization. First, to estimate individual differences in gradients’ spatial layout, trend surface modelling (TSM) was applied to each subject-level connectopic map30,57,64. This spatial-statistics approach parameterizes gradients at subject level, yielding a set of model parameters describing the topographic characteristics of each gradient in x, y, z directions (see Methods, and Supplementary Figure 5 for model selection). Unlike voxel-wise statistical inference on gradients, which overlooks the high interdependence between voxels’ gradient values and demands rigorous correction for multiple comparisons, TSM allows for statistical inference across a region’s internal heterogeneity using a concise set of independent parameters64. Moreover, by adjusting the number of polynomial terms, TSM facilitates examination of spatial trends across gradients at coarser-to-finer levels64.

Individual maps of D1DR binding potential (BP) were also submitted to TSM, yielding a set of spatial model parameters describing the topographic characteristics of hippocampal D1DR distribution for each participant. D1DR parameters were subsequently used as predictors of gradient parameters in one multivariate GLM per gradient (in total 6 GLMs, controlled for age, sex, and mean FD). Results are reported with p-values at an uncorrected statistical threshold and p-values after adjustment for multiple comparisons using the Benjamini-Hochberg method to control the false discovery rate (FDR). Individual differences in D1DR topography significantly explained topography of right-hemisphere G2 (F = 1.207, p = 0.041, pFDR = 0.073; partial η2 = 0.118), but not of G1 nor G3 (F 0.953-1.108, p 0.222-0.596) (Figure 3). This association was robust across multiple TSM model orders (Supplementary Figure 6). Complimentary analyses were then conducted to further evaluate G2 as a dopaminergic hippocampal mode by utilizing additional DA markers at group-level. First, a map of D1DR distribution was formed by averaging the [11C]SCH23390 BP images across DyNAMiC participants (n=176), and a map of D2DR distribution was formed by averaging [11C]raclopride BP images in a subsample of DyNAMiC participants (n = 20). Previously published maps of DAT65 and FDOPA (https://www.nitrc.org/projects/spmtemplates) were also analyzed. Correlations across group-level TSM parameters66, revealed significant positive associations between G2 and D1, DAT, and FDOPA (D1: r = 0.501, p < 0.01, pFDR = 0.021; DAT: r = 0.378, p < 0.01, pFDR = 0.021; FDOPA: r = 0.584, p < 0.01, pFDR = 0.021; Figure 3C), although not D2 (D2: r = 0.131, p = 0.440, pFDR = 0.528), whereas correlations were not significant for G1 or G3, indicating that G2 best captures shared principles of functional and molecular organization.

Shared functional and molecular organization within the hippocampus.

A) Multivariate effects of hippocampal D1DR TSM parameters as predictors of G2 TSM parameters. Images show average organization of G2 and D1DR in the right hemisphere. Note that the arbitrary color scale of G2 has been flipped. B) Correlations between group-level TSM parameters of functional gradients and DA markers.

Dedifferentiated gradient topography in older age

Effects of age on gradient topography were assessed using multivariate GLMs including age as the predictor and gradient TSM parameters as dependent variables (controlling for sex and mean frame-wise displacement; FD). One model was fitted per gradient and hemisphere, each model including all TSM parameters belonging to a gradient (in total, 6 GLMs). There was a significant effect of age on topographic characteristics of all three gradients. G1 displayed the greatest effect of age (left: F(9,150) = 5.853, p < 0.001, pFDR = 0.003, partial ƞ2 = 0.260; right: F(9,150) = 6.971, p < 0.001, pFDR = 0.003, partial ƞ2 = 0.298), followed by G2 (left: F(12,147) = 2.583, p = 0.004, pFDR = 0.01, partial ƞ2 = 0.174; right: F(12,145) = 2.635, p = 0.003, pFDR = 0.008, partial ƞ2 = 0.179), and G3 (left: F(12,147) = 1.973, p = 0.030, pFDR = 0.056, partial ƞ2 = 0.139; right: F(12,145) = 2.082, p = 0.021, pFDR = 0.042, partial ƞ2 = 0.147). To visualize effects, subject-level G1 and G2 values were plotted along the anteroposterior axis, averaged within young (20-39 years), middle-aged (40-59 years) and older (60-79 years) adults. Connectivity across G1 and G2 displayed less distinct differentiation at older age, depicted by the flatter curves in the older group (Figure 4A). The difference in gradient values between each consecutive pair of gradient bins was computed, and the locations of significant univariate contributions to the effect of age on these values were visualized along gradients (Figure 4B-C). For G1, this primarily localized effects to a middle hippocampal region extending posteriorly from just after the uncal apex (MNI y=21).

Effects of age on hippocampal gradients.

A-B) Less specificity in connectivity change across G1 and G2 in older age. Average values of subject-level gradient bins are plotted as a function of their distance from the most anterior hippocampal voxel. Separate lines mark young (20-39 years; gray), middle-aged (40-59 years, green), and older (60-79 years; blue) age groups. The flatter curves in the older group indicate less distinct change in connectivity patterns across gradients in older age. C) Localization of age effects along G1. Shaded fields indicate the position of significant age effects along G1. D) Localization of age effects along G2. Shaded fields indicate the position of significant age effects along G2.

Topography of hippocampal gradients predicts episodic memory performance

Next, we tested associations between topography of the three gradients and episodic memory. Using hierarchical multiple regression models in which age, sex, and mean FD were controlled for in a first step (M1), we entered TSM parameters of the three gradients as predictors of episodic memory in a step-wise manner. Models were assessed separately for left and right hemispheres, across the full sample and within age groups, yielding eight hierarchical models in total. Results are reported with p-values at an uncorrected statistical threshold and p-values after FDR adjustment. Memory performance was, across the sample, predicted by G2 in the left hippocampus (Adj. R2 = 0.308, ΔR2 = 0.096, F= 1.842, p = 0.047, pFDR = 0.082) over and above covariates and topography of G1, which did not predict performance (Adj. R2 = 0.260, ΔR2 = 0.029, F= 0.695, p = 0.713, pFDR = 0.771), and neither did G3 (Adj. R2 = 0.276, ΔR2 = 0.027, F= 0.502, p = 0.910, pFDR = 0.920), Figure 5A. Observing that the association between G2 and memory did not remain significant after FDR adjustment, we performed the same analysis in our replication dataset, which also included episodic memory testing. Consistent with the observation in our main dataset, G2 significantly predicted memory performance (Adj. R2 = 0.368, ΔR2 = 0.081, F= 1.992, p = 0.028) over and above covariates and topography of G1. Here, the analysis also showed that G1 topography predicted performance across the sample (Adj. R2 = 0.325, ΔR2 = 0.112, F= 3.431, p < 0.001).

Gradient topography and episodic memory.

A) Individual differences in topographic characteristics of left-hemisphere G2 significantly predicted episodic memory performance across the sample, over and above the first- and second-step models (M1: age, sex, in-scanner motion; G1 parameters). B) Topographic characteristics of G1 in the left hemisphere significantly predicted episodic memory performance in young adults, over and above M1 (age, sex, and in-scanner motion).

In our main dataset, memory performance was within young adults, predicted by left-hemisphere G1 (Adj. R2 = 0.182, ΔR2 = 0.357, F= 2.672, p = 0.015, pFDR = 0.030), whereas neither G2 (Adj. R2 = 0.204, ΔR2 = 0.191, F= 1.098, p = 0.396, pFDR = 0.492) nor G3 (Adj. R2 = 0.384, ΔR2 = 0.236, F= 1.755, p = 0.132, pFDR = 0.189) improved the prediction (Figure 5B). No gradient predicted memory within middle-aged or older age groups (F 0.432-1.865, p 0.928-0.113; Supplementary Table 1).

Youth-like gradient topography supports memory in older age

To investigate the functional role of G1 in old age, the principal and most age-sensitive gradient, we tested whether memory in older adults would be facilitated by youth-like gradient topography using two independent datasets. In DyNAMiC, we applied data-driven latent class analysis (LCA) to TSM parameters (residualized to account for age, sex, and mean FD) of left-hemisphere G1, which predicted episodic memory performance in younger adults. LCA yielded a two-class solution identifying two subgroups (n=19 vs. n=30) of older adults (60-79 years), which by definition differed in left-hemisphere G1 characteristics (F(9,37) = 13.778, p < 0.001, pFDR = 0.003, partial η2 = 0.770). A difference was also evident in the right hemisphere (F(9,37) = 3.790, p = 0.002, pFDR = 0.005, partial η2 = 0.480).

Individuals in the smaller subgroup were determined as exhibiting an aged gradient profile, whereas older adults in the larger subgroup as exhibiting a youth-like gradient profile. The classification based on G1 parameters extended across all three gradients in both hemispheres (Figure 6A), such that the smaller subgroup displayed marked differences from younger adults across all gradients (left G1: F(9,63) = 15.549, p < 0.001, pFDR = 0.003, partial η2 = 0.690; right G1: F(9,63) = 5.322, p < 0.001, pFDR = 0.003, partial η2 = 0.432; left G2: F(12,60) = 3.991, p < 0.001, pFDR = 0.003, partial η2 = 0.444; right G2: F(12,60) = 2.192, p = 0.023, pFDR = 0.045, partial η2 = 0.305; left G3: F(12,60) = 2.832, p = 0.004, pFDR = 0.01, partial η2 = 0.362; right G3: F(12,60) = 1.844, p = 0.061, pFDR = 0.098, partial η2 = 0.269), while the larger subgroup differed less from young adults in terms of G1 (left G1: F(9,74) = 4.416, p < 0.001, pFDR = 0.003, partial η2 = 0.349; right G1: F(9,74) = 3.086, p = 0.003, pFDR = 0.008, partial η2 = 0.273), and displayed second- and third-order gradients comparable to those in younger age (left G2: F(12,71) = 1.616, p = 0.107, pFDR = 0.167, partial η2 = 0.215; right G2: F(12,71) = 1.442, p = 0.168, pFDR = 0.235, partial η2 = 0.196; left G3: F(12,71) = 1.122, p = 0.357, pFDR = 0.457, partial η2 = 0.159; right G3: F(12,71) = 1.596, p = 0.112, pFDR = 0.172, partial η2 = 0.212). See Supplementary Figure 7 for classification based on right-hemisphere G1.

Gradient profiles in older age.

A) Two older subgroups were identified based on left-hemisphere G1. The first group (n=19) displayed gradient characteristics significantly different from those in young adults, whereas the second group (n=30) displayed gradient characteristics more similar to those in young adults. Bars represent comparisons of gradient TSM parameters between older subgroups and younger adults. B) Average gradient values across participants within subgroups, plotted against the distance (in mm) from the most anterior hippocampal voxel. The flatter curves in older adults with an aged gradient profile suggest less distinct change in connectivity across gradients. C) Group-level G1 and G2 for young, youth-like, and aged groups. D) Position of cortical networks in gradient space. E) Older subgroups were comparable in terms of age, sex, hippocampal gray matter volume, and average levels of hippocampal D1DR availability, while older adults with a youth-like gradient profile exhibited a more youth-like profile also in D1DR distribution, and superior episodic memory.

Plotting connectivity change along G1 and G2 in the aged and youth-like subgroups revealed that the diminished topographic specificity observed across gradients in older individuals (Figure 4A) was driven by older adults with an aged gradient profile (Figure 6B). Both older subgroups displayed altered gradient organization across cortex (Figure 6C-D). The distribution of cortical networks in G1 space indicated a unimodal-transmodal organization in youth-like older adults, not evident in the aged older group (Figure 6D). The two groups did not differ in terms of age (aged: 70.8±6.0; youth-like: 68.4±4.7; t = 1.548, p = 0.128, pFDR = 0.189), sex (aged: 9 men/10 women; youth-like: 16 men/14 women; X2 = 0.166, p = 0.684, pFDR = 0.746), nor hippocampal gray matter (left hemisphere: aged: 4271.2 ml ±480.9; youth-like: 4246.8 ml ±269.1; t = 0.223, p = 0.824, pFDR = 0.850; right hemisphere: aged: 3866.2 ml ±446.3; youth-like: 3979.9 ml ±398.1; t = 0.929, p = 0.357, pFDR = 0.457). Subgroups showed comparable levels of average hippocampal D1DR availability (left: youth-like 0.257±0.06; aged 0.242±0.06; t = 0.796, p = 0.430, pFDR = 0.525; right: youth-like 0.242±0.06; aged 0.251±0.06 t = 0.296, p = 0.768, pFDR = 0.817), but displayed a pattern of differences in D1DR TSM parameters in comparison to young adults supporting youth-like and aged profiles (youth-like subgroup vs. young: left: F(9,74) = 1.645, p = 0.118, pFDR = 0.176, partial η2 = 0.167; aged subgroup vs. young: left: F(9,62) = 3.478, p = 0.002, pFDR = 0.005, partial η2 = 0.335; Figure 6E).

In line with our hypothesis, we observed superior memory in older adults exhibiting a youth-like gradient profile (Figure 6E): at trend-level for the composite episodic measure (aged: 43.2±3.7; youth-like: 46.5±6.6; t = 1.958, p = 0.056, pFDR = 0.092), driven by a significant group difference on its word recall sub test (aged: 40.9±4.5; youth-like: 43.4±6.8; t = 2.600, p = 0.012, pFDR = 0.025). Word recall performance was furthermore predicted by left-hemisphere G1 parameters (over and above age, sex, and mean FD) in the youth-like older adults (Adj. R2 = 0.464, ΔR2 = 0.543, F = 3.043, p = 0.028, pFDR = 0.054), while no association was observed in the aged older subgroup (Adj. R2 = 0.063, ΔR2 = 0.533, F = 1.004, p = 0.518, pFDR = 0.599). Crucially, the ability of left G1 topography to inform classification of older adults into mnemonically distinct subgroups was replicated in the Betula sample (Supplementary Information; Figure 7C).

Discussion

In this study, we present a comprehensive multidimensional characterization of functional hippocampal-neocortical integration across the adult human lifespan, and methodically map topography of connectivity gradients onto behavioral and molecular phenotypes. Our findings reveal the presence of two distinct connectivity gradients distributed along the hippocampus longitudinal axis which contribute to episodic memory function across the lifespan. These observations underscore the importance of disentangling multiple dimensions of hippocampal functional organization in advancing our understanding of cortico-hippocampal systems for memory-related behavior. Moreover, we demonstrate in two independent samples, that maintaining a youth-like gradient profile in older age – characterized by preserved distinctiveness of connectivity change along gradients – supports mnemonic functioning, and that increased homogeneity of gradient topography may precede gray matter atrophy.

Despite converging evidence in young adults indicating multiple overlapping modes of functional connectivity across the hippocampus21,29–32, significant questions have remained regarding their spatial distribution across cortex, and their role in behavior. Here, connectopic mapping57 identified a principal anteroposterior gradient (G1); a second-order gradient of mid-to-anterior/posterior long-axis variation (G2); and a third-order gradient conveying variation across the hippocampus transverse axis (G3). Although we restrict our discussion of G3 given its low proportion of explained variance, we note that it mirrors patterns previously observed in both structure and function13,31,67. Consistently, cortical patterns of G3 matched cortical connectivity profiles previously reported for hippocampal subfields29,62. Our results confirm that functional data favors detection of anteroposterior hippocampal organization in contrast to that determined by its cytoarchitecture21, but indicate that higher-order connectivity modes may indeed carry coarse-scale information about subfield-determined organization. Greater anatomical specificity, with more precise characterization of connectivity in relation to subfield boundaries while minimizing effects of inter-individual differences in hippocampal shape and folding, might be achieved by adopting techniques implementing a geodesic coordinate system to represent effects within the hippocampus68,69.

The general organization of neocortical connectivity within the hippocampus showed stability across the lifespan (Figure 1), although clear effects of age were evident in the fine-scale topography of connectivity modes (Figure 4). Older age was associated with less distinct transitions in connectivity along G1 and G2, an effect that was exacerbated in a subgroup of older adults exhibiting an aged gradient profile – separated from older adults exhibiting youth-like gradient topography (Figure 6A-B). This finding constitutes an important addition to evidence highlighting increased homogeneity in hippocampal function in aging44,70. Current theories view this loss of specificity as a consequence of its functional isolation from neocortical areas, possibly linked to tau-driven degeneration of the perforant pathway44,71. Age-related deterioration of this entorhinal-hippocampal pathway has in turn been linked to impaired mnemonic functioning72. Importantly, older individuals exhibiting dedifferentiated gradient topography, but comparable hippocampal volumes, displayed less efficient episodic memory compared to older adults maintaining youth-like gradient topography (Figure 6; Supplementary Figure 7). This underscores the potential of gradient-based techniques to capture behaviorally relevant alterations in hippocampal function at a stage preceding structural decline.

Testing the correspondence across cortex of G1 and G2 to canonical gradients of cortical function provided support for G1 reflecting differentiation along a representation-modulation dimension (e.g., task-negative/task-positive35; Figure 1C-E), separating frontoparietal areas of attention and control from sensorimotor and DMN areas. Consistent with age-related loss of segregation between task-positive and task-negative poles45, also evident in numerous diseases involving hippocampal dysfunction (e.g., depression73, schizophrenia74, and Alzheimer’s disease46), we observed altered cortical organization of G1 in both older subgroups. A main difference between subgroups, however, included a unimodal-transmodal organization instead emerging in youth-like older adults, whereas not in the aged subgroup (Figure 6D). A meaningful role of this potential reorganization was indicated by an association between G1 topography and memory in the youth-like older adults only. In terms of G2, we provide support of its characterization as a local representation of the principal unimodal-transmodal cortical gradient (Figure 1C-E), widely demonstrated across functional, structural, and molecular modalities33,34,75. Taken together, our observations support a framework of cortico-hippocampal integration in which the DMN is positioned in different neural contexts: in one case at the opposite end from frontoparietal networks of attention and control, and in the other, opposite to sensorimotor and visual networks – indicating that an account of the hippocampus functional connectivity with the DMN is dependent on multiple neurofunctional hierarchies. The overlap of G1 and G2 may potentially reflect the superimposition of gradients and hubs of long-axis anatomical connections indicated in both the human and animal hippocampus20,76. Moreover, the observation that macroscale relationships between distinct cortical systems are mapped out by G1 and G2 in this manner may reflect the hippocampus primordial position in the laminar development of the cerebral cortex77, supporting the idea that hippocampal function stands, from a phylogenetic perspective, to inform general principles of brain organization21.

We discovered that G2, specifically, manifested organizational principles shared among function, behavior, and neuromodulation. Meta-analytical decoding reproduced a unimodal-associative axis across G2 (Figure 3B), and analyses in relation to the distribution of D1DRs – which vary across cortex along a unimodal-transmodal axis78,79 – demonstrated topographic correspondence both at the level of individual differences and across the group. It should, however, be acknowledged that PET imaging in the hippocampus is associated with resolution-related limitations, although previous research indicate high test-retest reliability of [11C]SCH23390 PET to quantify D1DR availability in this region80. As such, mapping the distribution of hippocampal D1DRs at a fine spatial scale remains challenging, and replication of our results in terms of overlap with G2 is needed in independent samples. Here, we evaluated the observed spatial overlap between G2 topography and D1DRs across multiple TSM model orders, showing correspondence between modalities from simple to more complex parameterizations of their spatial properties. Topographic correspondence was additionally observed between G2 and other DA markers from independent datasets (Figure 3B), suggesting that G2 may constitute a mode reflecting a dopaminergic phenotype, which contributes to the currently limited understanding of its biological underpinnings.

Results linked both G1 and G2 to episodic memory, suggesting complimentary contributions of these two overlapping long-axis modes. Considered together, analyses in the main and replication datasets indicated a role of G2 topography in memory across the adult lifespan, independent of age. A similar association with G1 was only evident across the entire sample in the replication dataset, whereas results in the main sample seemed to emphasize a role of youth-like G1 topography in memory performance. In line with previous research, memory was successfully predicted by G1 topography in young adults30, and similarly predicted by G1 in older adults exhibiting a youth-like functional profile.

It is noteworthy that meta-analytical decoding of G2 primarily linked the unimodal connectivity patterns of anterior and posterior subregions to terms of episodic memory, encoding, and navigation (Figure 2C). G2’s role in memory might as such be considered in light of hippocampal integration with the visual system, by which it contributes to complex perceptual processes supporting memory81. In humans, there is evidence of direct hippocampal connections to early visual areas, with recent tractography-based work demonstrating connectivity as primarily localized to the posterior hippocampus and to a smaller region in the anterior hippocampus76.

The verbal nature of our memory tasks likely contributed to the left-lateralization of effects, yet, predominant left-hemisphere vulnerability to aging and age-related pathology should not be ruled out as a meaningful contributor to these effects82. Average hippocampal D1DR availability did not differ between older subgroups, but a tendency towards youth-like and aged functional profiles being mirrored in D1DR topography was observed (Figure 6E). However, longitudinal data is ultimately required to inform the underlying mechanisms of individual differences in hippocampal gradient topography in older age83,84. Future studies should, furthermore, assess gradients’ modulation by behavioral conditions and extend these methods to clinical samples characterized by hippocampal dysfunction.

This study establishes behavioral relevance of two overlapping long-axis modes of hippocampal-neocortical functional connectivity, shedding light on their age-related dedifferentiation, and its impact on cognition. In sum, this study introduces a multidimensional framework for understanding hippocampal-neocortical integration and its interplay with memory and neuromodulation throughout the adult human lifespan.

Materials and Methods

This study included data from the DopamiNe, Age, connectoMe, and Cognition (DyNAMiC) study, for which the design and procedures have been described in detail elsewhere58,85. Here, we include the materials and methods relevant to the current study. DyNAMiC was approved by the Regional Ethical board and the local Radiation Safety Committee of Umeå, Sweden. All participants provided written informed consent prior to testing.

Participants

The DyNAMiC sample included 180 participants (20-79 years; mean age = 49.8±17.4; 90 men/90 women equally distributed within each decade). Individuals were randomly selected from the population register of Umeå, Sweden, and recruited via postal mail. Exclusion criteria implemented during the recruitment procedure included brain pathology, impaired cognitive functioning (Mini Mental State Examination < 26), medical conditions and treatment that could affect brain functioning and cognition (e.g. dementia, diabetes, and psychiatric diagnosis), and brain imaging contraindications (e.g. metal implants). All participants were native Swedish speakers. A total of 16 participants were excluded from connectopic mapping due to excessive in-scanner motion, leaving resting-state fMRI data for 164 participants (20-78 years; mean age = 48.7±17.3). As a replication data set, we used an independent sample of 224 cognitively healthy and native Swedish-speaking adults (122 men/102 women; 29-85 years mean age = 65.0±13.0) from the population-based Betula project, for which the design and recruitment procedures have been reported in detail elsewhere59.

Episodic memory

Episodic memory was measured using three tasks testing word recall, number-word recall and object-location recall, respectively58. In the word recall task, participants were presented with 16 Swedish concrete nouns that appeared successively on a computer screen. Each word was presented for 6 s during encoding with an inter-stimulus interval (ISI) of 1 s. Following encoding, participants reported as many words as they could recall by typing them using the keyboard. Two trials were completed, yielding a maximum score of 32. In the number-word task, participants encoded pairs of 2-digit numbers and concrete plural nouns (e.g., 46 dogs). During encoding, eight number-word pairs were presented, each displayed for 6 s, with an ISI of 1 s. Following encoding, nouns were presented again, in a re-arranged order, and participants had to report the 2-digit number associated with each presented noun (e.g. How many dogs?). This task included two trials with a total maximum score of 16. The third task was an object-location memory task. Here, participants were presented with a 6 × 6 square grid in which 12 objects were, one by one, shown at distinct locations. Each object-position pairing was displayed for 8 s, with an ISI of 1 s. Following encoding, all objects were simultaneously shown next to the grid for the participant to move them (in any order) to their correct position in the grid. If unable to recall the correct position of an object, participants had to guess and place the object in the grid to the best of their ability. Two trials of this task were completed, making the total maximum score 24.

A composite score of performances across the three tasks was calculated and used as the measure of episodic memory. For each of the three tasks, scores were summarized across the total number of trials. The three resulting sum scores were z-standardized and averaged to form one composite score of episodic memory performance (T score: mean = 50; SD = 10). Missing values were replaced by the average of the available observed scores.

Image acquisition

Brain imaging was conducted at Umeå University Hospital, Sweden. Structural and functional MRI data were acquired with a 3T Discovery MR 750 scanner (General Electric, WI, USA), using a 32-channel head coil. Positron emission tomography (PET) data were acquired with a Discovery PET/CT 690 scanner (General Electric, WI, USA).

Structural MR Imaging

Anatomical T1-weighted images were acquired with a 3D fast-spoiled gradient-echo sequence, collected as 176 slices with a thickness of 1 mm. Repetition time (TR) was 8.2 ms, echo-time (TE) = 3.2 ms, flip angle = 12°, and field of view (FOV) = 250 × 250 mm.

Functional MR Imaging

Functional MR data were collected during resting-state, with participants instructed to keep their eyes open and focus on a fixation cross during scanning. Images were acquired using a T2*-weighted single-shot echo-planar imaging (EPI) sequence, with a total of 350 volumes collected over 12 minutes. The functional time series was sampled with 37 transaxial slices, slice thickness = 3.4 mm, and 0.5 mm spacing, TR = 2000 ms, TE = 30 ms, flip angle = 80°, and FOV = 250 x 250 mm. Ten dummy scans were collected at the start of the sequence.

PET Imaging

PET was conducted in 3D mode with a Discovery PET/CT 690 (General Electric, WI, US) to assess whole-brain dopamine D1 receptor availability using the radioligand [11C]SCH23390. Scanning was done during a resting condition, with participants instructed to lay still and remain awake with their eyes open. To minimize head movement, a thermoplastic mask (Posicast®; CIVCO medical solutions; IA, US) was individually fitted for each participant, and attached to the bed surface during scanning. Following a low-dose CT scan (10 mA, 120 kV, and 0.8 s rotation time) for attenuation correction, an intravenous bolus injection with target radioactivity of 350 MBq [11C]SCH23390 was administered. The PET scan was a 60 min dynamic scan, with 6 x 10 s, 6 x 20 s, 6 x 40 s, 9 x 60 s, 22 x 120 s frames. The average radioactivity dose administered to participants was 337±27 MBq (range 205-391 MBq). Due to participant drop-out and technical issues, complete PET data was available for 177 DyNAMiC participants.

Image preprocessing

Hippocampal segmentation and volumetric assessment

Individual anatomical T1-weighted images were submitted to automated segmentation in FreeSurfer version 686. A mean image of participants’ normalized T1-weighted images was also segmented in FreeSurfer, and yielded hippocampal and cortical segmentations used as masks for connectopic mapping. Regional gray matter (GM) volume was estimated from subject-specific hippocampal segmentations, and were corrected for total intracranial volume (ICV; the sum of volumes for grey matter, white matter, and cerebrospinal fluid). Adjusted volumes were equal to the raw volume - b(ICV - mean ICV), where b is the regression slope of volume on ICV87. Automated segmentation of the hippocampus into subiculum, CA1-3, and DG/CA4 subfields was conducted in FreeSurfer using the group-average T1-weighted image, for sample-specific masks to overlay onto G3 (Supplementary Figure 2).

Functional MRI data

Resting-state fMRI data were preprocessed using Statistical Parametric Mapping (SPM12: Wellcome Trust Centre for Neuroimaging, http://www.fil.ion.ucl.ac.uk/spm/) implemented in an in-house software, DataZ. Functional images were slice-timing corrected, co-registered to the anatomical T1-images, and motion corrected, and underwent distortion correction using subject-specific B0-field maps. The functional data were subsequently co-registered to the anatomical T1-images again, temporally demeaned and linear and quadratic effects were removed. Next, a 36-parameter nuisance regression model was applied88, including mean cerebrospinal, white-matter, and whole-brain signal in addition to six motion parameters, including parameters’ squares, derivatives, and squared derivatives. To further control for in-scanner motion, the model also included a set of spike regressors, defined as binary vectors of motion-contaminated volumes exceeding a volume-to-volume root-mean-squared (RMS) displacement of 0.25 mm. A temporal high-pass filter (with a threshold of 0.009 Hz) was applied simultaneously as nuisance regression in order to not re-introduce nuisance signals. Finally, images were normalized to MNI space by Diffeomorphic Anatomical Registration using Exponentiated Lie algebra (DARTEL89) and smoothed with a 6-mm FWHM Gaussian kernel. Four individuals were excluded from the template-generation step due to non-pathological anatomical irregularities. In total, 16 participants were excluded due to displaying excessive in-scanner motion, as defined by displaying i) more than 20 volumes with >0.25 relative RMS difference in motion, and ii) greater than 0.2 average RMS across the run. On average, the relative RMS difference in motion across the sample was 0.090 (± 0.063), and the mean frame-wise displacement (FD) was 0.164 (± 0.104).

Dopamine D1 receptor availability

Preprocessing of PET data was performed in SPM12 (Wellcome Trust Centre for Neuroimaging, http://www.fil.ion.ucl.ac.uk/spm/). Binding potential relative to non-displaceable binding in a reference region (BPND; Innis et al., 2007), was used as an estimate of receptor availability (i.e. D1DR) in the hippocampus, for each participant defined using the FreeSurfer segmentation of their anatomical images. Cerebellum was used as reference region. PET images were corrected for head movement by using frame-to-frame image co-registration, and co-registered with T1-weighted MRI images with re-slicing to T1 voxel size. The simplified reference-tissue model (SRTM) was used to model regional time-activity course (TAC) data. Regional TAC data were adjusted for partial volume effects (PVE) by using the symmetric geometric transfer matrix (SGTM) method implemented in FreeSurfer, and an estimated point-spread-function of 2.5 mm full-width-at-half-maximum (FWHM). We additionally used data from a publicly available database of group-averaged volumetric maps of molecular target distributions (https://github.com/netneurolab/hansen_receptors). Specifically, we downloaded previously published maps of DAT65 and FDOPA (https://www.nitrc.org/projects/spmtemplates), to investigate the spatial correspondence between functional gradients and dopaminergic target distributions.

Mapping gradients of functional connectivity

Connectopic mapping57 was run through the ConGrads toolbox57 implemented in FSL91. Mapping was conducted on both subject level and group level, for the left and right hippocampus separately, and involved two main steps. First, for every hippocampal voxel, connectivity fingerprints were computed as the Pearson correlation between the voxel-wise time-series and a singular-value decomposition (SVD) representation of all cortical voxels. In a second step, non-linear manifold learning (Laplacian eigenmaps) was applied to a matrix expressing the degree of similarity between the voxel-wise fingerprints. This yields eigenvectors, so called connectopic maps, representing modes of functional connectivity (i.e. functional gradients). Each connectopic map is then projected onto cortex, for which each vertex is color coded according to the voxel in the hippocampus it correlates the most with. Since connectopic mapping at group level involves applying Laplacian eigenmaps to a group-average similarity matrix, group level mapping across the sample was conducted using the hippocampal and cortical masks derived from the FreeSurfer segmentation of a sample-mean structural image. Mapping was specified to compute 20 gradients, and a subsequent scree plot over explained variance indicated meaningful contributions of the three first connectopic maps, together explaining 67% of the variance across hemispheres (Supplementary Figure 1).

Stability of functional gradients across levels of spatial smoothing

It has recently been suggested that the reliability of connectopic mapping to detect meaningful gradients of resting-state functional connectivity may be limited due to spatial smoothing implemented during preprocessing of fMRI data60. To determine the stability of our hippocampal gradients, we conducted a series of control analyses across varying levels of smoothing, and in contrast to connectopic maps derived through connectopic mapping on random data. These analyses are presented in the Supplementary Information, and confirmed high stability of resting-state gradients and their ability to capture inter-individual differences, whereas random data failed to produce meaningful gradients (Supplementary Figure 4).

Alignment of connectopic maps across participants

To ensure optimal alignment of identified connectopic maps across participants, we employed Procrustes alignment, based on voxel-wise correlations, to order subject-level connectopic maps according to their correspondence with a set of reference maps (i.e. gradients computed at group level across the full sample). Moreover, whereas the sign of connectopic maps is arbitrary, differences therein have an impact on the spatial model parameters describing the topographic characteristics of gradients, derived through TSM in a later step. As such, the sign of subject-level connectopic maps showing negative correlations with the corresponding group-level reference map were inversed.

Trend surface modelling

Using spatial statistics, the topography of a connectopic map can be represented by a small number of spatial model parameters. This parameterization enables analyses of inter-individual differences, and is achieved through trend surface modelling (TSM), implemented in a third step of the ConGrads analysis pipeline57. In this step, the spatial pattern of each subject-level connectopic map is approximated by estimating a spatial statistical model. Model estimation involves fitting a set of polynomial basis functions along canonical axes of the connectopic map. In MNI space, this entails estimation along x, y, and z axes of the hippocampus. Thus, fitting a polynomial of degree 1 yields three TSM parameters (x, y, z), with any increase in model order corresponding to an increase in number of parameters (e.g. 6 parameters for the second model order: x, y, z, x2, y2, z2; 9 parameters for the third model order, etc.). Trend surface models are fitted with an increasing polynomial degree using Bayesian linear regression, which provides likelihood estimates that can be used for subsequent model selection. Here, model selection was based on three information sources: a) the Bayesian Information Criterion (BIC) across subjects for models estimated at orders 1-10; b) the % explained variance in connectopic maps by each model; and c) visual inspection of group-level gradients reconstructed from TSM parameters at different model orders. The purpose of using multiple information sources, as opposed to simply BIC, was to find a trade-off between high-quality reconstructions of gradients by TSM models, while keeping the number of model parameters sufficiently low for multivariate statistical analyses. A model order of 3 (=9 TSM parameters) was selected for G1, whereas a model order of 4 (=12 TSM parameters) was selected for G2 and G3 (Supplementary Figure 5). Each gradient’s set of TSM parameters were then used as either dependent or independent variables in multivariate GLMs investigating links between gradient topography and variables such as age, episodic memory performance, and D1DR distribution.

Transitions in connectivity as a function of the hippocampal longitudinal axis

To visualize the orthogonal patterns of change in connectivity conveyed by each gradient, and to aid in the interpretation of age effects, we divided each subject-level connectopic map into 23 bins of ∼2mm along the hippocampus anterior-posterior axis and estimated the average gradient value (ranging from 0-1) for each bin. Plotting the values of each bin against their distance in mm from the most anterior voxel in the hippocampus as such demonstrates the pattern of change in connectivity along the anterior-posterior axis30.

Cortical projections and correlations with gradients of cortical function

In ConGrads57, cortical mapping is computed independently for each gradient based on the regression equation: where A corresponds to data inside (i.e., hippocampus) and B to data outside (i.e., cortical) the ROI, and X-1 is the pseudo-inverse of the corresponding eigenvector (stacked with a row of ones). For each hippocampal gradient, the volumetric MNI-registered group-level cortical projection map was resampled to the Human Connectome Project 32k_LR midthickness surface mesh, using the volume-to-surface algorithm with enclosing mapping available in Connectome Workbench v.1.5.0. Spearman correlations were computed between the cortical projection of each hippocampal gradient and the three gradients previously established as the main axes of cortical functional organization34. To reduce the degree of freedom, each surface projection was resampled by 1000-cortical parcels92. Statistical significance of correlations was assessed by spin-test permutation93, randomly rotating a spherical projection of the cortical maps 1000 times, with two-tailed statistical significance determined at a 95% confidence level.

Mapping behavioral transitions along gradients using Neurosynth

Transitions in behavioral domains were mapped onto G1 and G2 using meta-analytical decoding in Neurosynth63. We assessed two sets of behavioral terms (Figure 2), the first was a selection of terms commonly linked to anteroposterior hippocampal functional specialization13,14, and the second a selection of terms based on a previous report demonstrating behavioral transitions along a unimodal-transmodal cortical axis34. For correspondence with meta-analytical maps, we created region of interest masks by projecting the cortical surface of each gradient to the 2-mm volumetric MNI152 standard space. These volumetric images were then divided into five twenty-percentile bins and binarized. The resulting images were used as input to the Neurosynth decoder, yielding an r statistic associated with each behavioral term per section of each gradient.